---
title: "Correct boundary of districts 13 and 15 in California from the 73rd to the 77th"
author: "Jeff Lewis"
date: 2025-11-13
format:
  html:
    self-contained: true
editor: source
---

```{r}
#| include: false
library(tidyverse)
library(sf)
library(ggthemes)
library(maptiles)
library(tidyterra)
library(leaflet)
```

## Summary

Ron Trigueiro (<a href="mailto:sansimeonfan\@aol.com">sansimeonfan\@aol.com</a>) wrote:

>Dear Dr. Lewis:
> 
>I'm engaged in a research project in which I've found your U.S. Congressional District Shapefiles useful. However, I wish to call >your attention to what I believe is an error in one of your maps: California_73_to_77.geojson.
> 
>Taking into account the loss of the district information--an issue you apparently have been dealing with since 2022--I'm less than> >100% sure that I'm describing this problem correctly, but I hope you'll bear with me.
>
>The issue is a boundary between congressional districts 13 and 15, in Los Angeles. To approach the problem, let's begin at the >junction of Los Feliz Blvd. and N. Western Ave., where the district boundary [District 15 to the west, District 13 to the east] >extends south along N. Western Ave. to its junction with Beverly Blvd. At this point, you show the boundary turning west along >Beverly Blvd. to its junction with N. Rossmore Ave. [where it meets the boundary with CD 16]. This is, I believe, the location of >the map error.
> 
>A less detailed map of this districting I've consulted indicates that if the line connecting Western and Rossmore aves. along >Beverly Blvd. is correct, District 15 is cut off and exists only north of Beverly Blvd. while its territory to the south is >incorrectly shown as being part of District 13. But the problem is solved if, in fact, the stretch of the boundary along Beverly >Blvd. extends east from N. Western Ave., NOT west.
> 
>I believe that the boundary between CDs 13 and 15 should be shown as turning east along Beverly Blvd. at N. Western Ave. and >extending initially to the junction of Beverly Blvd. [where that street turns southeast] and W. Temple St. [which proceeds, more or >less, straight ahead]. Beyond this point, the boundary apparently continues along Beverly Blvd. to its junction with N./S. Hoover >St. [where it meets the boundary with CD 14]. This change allows CD 15 to extend well south of Beverly Blvd.--as it apparently >did--and shows CD 13 as only extending a short distance down CD 14's western boundary, instead of practically engulfing the latter.
>
>I'm not the most technically astute person, so I hope this detailed description will enable you to investigate my query using >methods easily available to you. Please let me know if I was correct about the existence of this error and if so, if my suggestion >indeed is the solution.
> 
>Yours truly,
>Ron Trigueiro


Here are the districts in question.  A quick comparison of the map to the district definitions 
suggests that Congressional District 15 is supposed to contain Assembly District 58, but it appears 
that AD 58 has been placed in CD 13.  And, there may be other issues in the Los Angeles Couny districts.

```{r}
#| message: false
shp_ca <- st_read("California_073_to_077.geojson") 
cf <- colorFactor(topo.colors(3), domain = c(13, 14, 15))
m <- leaflet() |>
      addProviderTiles("CartoDB.Positron") |> 
      addPolygons(data = shp_ca |> 
                          filter(district %in% c(13,14,15)),
                  fillColor = ~cf(district),
                  color = NA,
                  fillOpacity = 0.2,
                  label = ~district) 
m
```


```{r}
#| message: false
shp_ca <- st_read("California_073_to_077_edit.json") 
cf <- colorFactor(topo.colors(3), domain = c(13, 14, 15))
m <- leaflet() |>
      addProviderTiles("CartoDB.Positron") |> 
      addPolygons(data = shp_ca |> 
                          filter(district %in% c(13,14,15)),
                  fillColor = ~cf(district),
                  color = NA,
                  fillOpacity = 0.2,
                  label = ~district) 
m
```
