---
title: "Fix Washington CD 4 Overlaps From Issue #18"
format: html
editor: source
---

## Overview

[Issue #18](https://github.com/JeffreyBLewis/congressional-district-boundaries/issues/18)
reports that in `Washington_099_to_102.geojson` (99th-102nd Congress,
1985-1993), District 4's polygon overlaps four neighboring districts by
small but non-zero amounts:

| Pair | Intersection (deg²) | ~Sq mi |
|---|---|---|
| WA-02 ∩ WA-04 | 0.000846 | ~2.7 |
| WA-03 ∩ WA-04 | 0.000144 | ~0.5 |
| WA-04 ∩ WA-05 | 0.000003 | ~0.01 |
| WA-04 ∩ WA-08 | 0.000014 | ~0.05 |

The reporter noted that District 4 is the common participant in every
overlap, suggesting District 4's own polygon is the source of the bad
edges rather than each neighbor independently. A follow-up comment from
the repo owner confirmed the overlaps are small relative to district size
and fall along water bodies (mainly the Columbia River), and flagged this
as part of a broader topology-cleanup effort. A later comment from a third
party additionally reported District 6 overlapping District 8 with a
non-contiguous piece of District 6; that's checked and addressed separately
below since it turns out not to be reproducible against the current file.

```{r}
#| message: false
library(tidyverse)
library(sf)
library(ggspatial)
sf_use_s2(FALSE)
```

## Confirming the bug

```{r}
GEOJSON_FILE <- "../../GeoJson/Washington_099_to_102.geojson"
wa <- st_read(GEOJSON_FILE, quiet = TRUE)

all_pairs <- combn(sort(unique(wa$district)), 2, simplify = FALSE)
overlaps <- map_dfr(all_pairs, function(p) {
  a <- wa |> filter(district == p[1]) |> st_geometry()
  b <- wa |> filter(district == p[2]) |> st_geometry()
  area <- as.numeric(st_area(st_intersection(a, b)))
  tibble(d1 = p[1], d2 = p[2], area_sqm = if (length(area) == 0) 0 else area)
}) |>
  filter(area_sqm > 0) |>
  arrange(desc(area_sqm))
overlaps
```

This reproduces exactly the four overlaps reported in the issue, and finds
no others anywhere else in the file -- confirming District 4 is indeed the
only district with a topology problem here.

## Determining the fix

Plotting one of the overlaps (District 2 vs. District 4, the largest) shows
the overlap isn't a small localized patch, but a thin ribbon running along
almost the *entire* shared boundary -- classic evidence that the two
districts' shared edge was digitized twice, as two slightly different
lines, rather than sharing a single common boundary:

```{r}
#| fig-width: 6
#| fig-height: 8
d2 <- wa |> filter(district == 2) |> st_geometry()
d4 <- wa |> filter(district == 4) |> st_geometry()
overlap_24 <- st_intersection(d2, d4)

ggplot() +
  geom_sf(data = d2, fill = NA, color = "red") +
  geom_sf(data = d4, fill = NA, color = "blue") +
  geom_sf(data = overlap_24, fill = "magenta", color = "magenta") +
  coord_sf(xlim = st_bbox(st_buffer(overlap_24, 0.05))[c("xmin", "xmax")],
           ylim = st_bbox(st_buffer(overlap_24, 0.05))[c("ymin", "ymax")]) +
  labs(title = "District 2 (red) / District 4 (blue), overlap in magenta") +
  theme_minimal()
```

Since District 4 is the only common participant, and none of its four
neighbors overlap *each other*, trimming District 4 back to exclude
whatever its neighbors already claim resolves every overlap at once without
touching District 2, 3, 5, or 8 at all -- consistent with the issue's own
diagnosis that District 4's polygon is the source of the bad edges. The
overlap slivers are tiny relative to District 4 (about 3.3 sq mi total, out
of a district of several thousand square miles), so there's no realistic
scenario where the "correct" fix instead requires enlarging District 4 at
its neighbors' expense.

## Fix

```{r}
d3 <- wa |> filter(district == 3) |> st_geometry()
d5 <- wa |> filter(district == 5) |> st_geometry()
d8 <- wa |> filter(district == 8) |> st_geometry()

neighbors <- st_union(c(d2, d3, d5, d8)) |> st_make_valid()
new_d4_geom <- st_difference(d4, neighbors) |> st_make_valid()

new_d4 <- wa |> filter(district == 4) |> st_set_geometry(new_d4_geom)
```

## Validation

The new District 4 is a valid MultiPolygon with the same six parts as
before (five untouched islands plus the trimmed mainland), no new interior
holes, and no overlap with any neighbor:

```{r}
st_is_valid(new_d4)

count_holes <- function(x) {
  sum(sapply(st_geometry(x), function(g) {
    parts <- if (inherits(g, "MULTIPOLYGON")) g else list(g)
    sum(sapply(parts, function(p) length(p) - 1))
  }))
}
count_holes(new_d4)

length(st_cast(st_geometry(new_d4), "MULTIPOLYGON", warn = FALSE)[[1]])

sapply(c(2, 3, 5, 8), function(d) {
  as.numeric(st_area(st_intersection(st_geometry(new_d4), wa |> filter(district == d) |> st_geometry())))
})
```

Area lost is negligible and matches the sum of the reported overlaps, and
every other district (including 2, 3, 5, and 8 themselves) is completely
untouched:

```{r}
as.numeric(st_area(d4)) - as.numeric(st_area(new_d4))

wa_other <- wa |> filter(district != 4)
new_wa <- bind_rows(wa_other, new_d4) |> arrange(district)

identical(
  wa |> filter(district != 4) |> st_geometry(),
  new_wa |> filter(district != 4) |> st_geometry()
)
```

Re-plotting the same District 2/4 boundary zoom from above, but with the
corrected District 4, shows the overlap ribbon is gone -- the blue line now
runs cleanly alongside the red one instead of straddling it:

```{r}
#| fig-width: 6
#| fig-height: 8
new_overlap_24 <- st_intersection(d2, st_geometry(new_d4))

ggplot() +
  geom_sf(data = d2, fill = NA, color = "red") +
  geom_sf(data = st_geometry(new_d4), fill = NA, color = "blue") +
  coord_sf(xlim = st_bbox(st_buffer(overlap_24, 0.05))[c("xmin", "xmax")],
           ylim = st_bbox(st_buffer(overlap_24, 0.05))[c("ymin", "ymax")]) +
  labs(title = "Corrected: District 2 (red) / District 4 (blue), no more overlap") +
  theme_minimal()
```

And the full corrected map, all eight districts:

```{r}
#| fig-width: 8
#| fig-height: 8
new_wa |>
  ggplot() +
  geom_sf(aes(fill = as.character(district))) +
  labs(title = "Washington, 99th-102nd Congress (corrected)", fill = "District") +
  theme_minimal()
```

## Visual validation against a real basemap

A closer look at the reported overlaps raised a natural follow-up: are
these overlaps/gaps small relative to real terrain features (roads,
rivers, lakes) the district lines should be following, or does the trimmed
edge look arbitrary? Plotting the corrected District 4 directly on an OSM
basemap answers that.

Along the Columbia River (District 3/4), the corrected boundary hugs the
river channel closely, including its small braids and islands -- exactly
matching the repo owner's original read that these boundaries follow real
water features:

```{r}
#| message: false
#| results: hide
#| fig-width: 9
#| fig-height: 7
d3_4326 <- st_transform(d3, 4326)
new_d4_4326 <- st_transform(st_geometry(new_d4), 4326)

bb_river <- st_bbox(c(xmin = -122.5, xmax = -121.7, ymin = 45.55, ymax = 45.8), crs = 4326)
ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  geom_sf(data = d3_4326, fill = NA, color = "red", linewidth = 1) +
  geom_sf(data = new_d4_4326, fill = NA, color = "blue", linewidth = 1) +
  coord_sf(xlim = bb_river[c("xmin", "xmax")], ylim = bb_river[c("ymin", "ymax")], crs = 4326) +
  labs(title = "District 3 (red) / corrected District 4 (blue) along the Columbia River") +
  theme_minimal()
```

Near the District 2/4 boundary in the remote North Cascades (around the
Chickamin Glacier, well above any road), the corrected line still runs
cleanly alongside District 2 with no jagged artifacts -- there's no
road/river to follow here, so the line most likely follows a section-line
or ridge, but critically the fix doesn't introduce any spurious zigzags:

```{r}
#| message: false
#| results: hide
#| fig-width: 9
#| fig-height: 9
d2_4326 <- st_transform(d2, 4326)

bb_glacier <- st_bbox(c(xmin = -121.3, xmax = -120.9, ymin = 48.25, ymax = 48.45), crs = 4326)
ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  geom_sf(data = d2_4326, fill = NA, color = "red", linewidth = 1) +
  geom_sf(data = new_d4_4326, fill = NA, color = "blue", linewidth = 1) +
  coord_sf(xlim = bb_glacier[c("xmin", "xmax")], ylim = bb_glacier[c("ymin", "ymax")], crs = 4326) +
  labs(title = "District 2 (red) / corrected District 4 (blue) near Chickamin Glacier") +
  theme_minimal()
```

(An earlier version of this patch tried to also close the small residual
gaps left along these same boundaries -- areas neither district claims,
the flip side of the overlaps fixed above -- using a snap/buffer-based
fill. That approach was dropped: it introduced exactly the kind of
spurious zigzag artifact this section checks for, in this same glacier
area, without any real feature to justify it. The gaps left by the
overlap-only fix here are thin, sub-visual slivers -- invisible at any
sane viewing scale -- so closing them isn't worth the risk of visibly
worse geometry.)

## The District 6 / District 8 report

A later comment on the issue reported District 6 overlapping District 8,
with the overlapping piece of District 6 non-contiguous from the rest of
District 6. Checking directly against the current file finds no overlap
between District 6 and District 8 at all:

```{r}
d6 <- wa |> filter(district == 6) |> st_geometry()
as.numeric(st_area(st_intersection(d6, d8)))
```

District 6 does have many disconnected parts (it's a Puget Sound district
with many islands, so some fragmentation is expected), but none of them
overlap District 8. This isn't reproducible against the current file, so
it's left out of this patch -- it may have already been resolved by
whatever produced the current file, or the report may have been against a
stale copy of the data. If it recurs, it likely needs its own issue/patch
rather than folding it in here.

## Write out the corrected file

```{r}
unlink(GEOJSON_FILE)
st_write(new_wa, GEOJSON_FILE, append = FALSE)
```
