---
title: "Fix Washington CD 4 Overlaps From Issue #18"
format: html
editor: source
---

## Overview

[Issue #18](https://github.com/JeffreyBLewis/congressional-district-boundaries/issues/18)
reports that in `Washington_099_to_102.geojson` (99th-102nd Congress,
1985-1993), District 4's polygon overlaps four neighboring districts by
small but non-zero amounts:

| Pair | Intersection (deg²) | ~Sq mi |
|---|---|---|
| WA-02 ∩ WA-04 | 0.000846 | ~2.7 |
| WA-03 ∩ WA-04 | 0.000144 | ~0.5 |
| WA-04 ∩ WA-05 | 0.000003 | ~0.01 |
| WA-04 ∩ WA-08 | 0.000014 | ~0.05 |

The reporter noted that District 4 is the common participant in every
overlap, suggesting District 4's own polygon is the source of the bad
edges rather than each neighbor independently. A follow-up comment from
the repo owner confirmed the overlaps are small relative to district size
and fall along water bodies (mainly the Columbia River), and flagged this
as part of a broader topology-cleanup effort. A later comment from a third
party additionally reported District 6 overlapping District 8 with a
non-contiguous piece of District 6; that's checked and addressed separately
below since it turns out not to be reproducible against the current file.

```{r}
#| message: false
library(tidyverse)
library(sf)
library(ggspatial)
sf_use_s2(FALSE)
```

## Confirming the bug

```{r}
GEOJSON_FILE <- "../../GeoJson/Washington_099_to_102.geojson"
wa <- st_read(GEOJSON_FILE, quiet = TRUE)

all_pairs <- combn(sort(unique(wa$district)), 2, simplify = FALSE)
overlaps <- map_dfr(all_pairs, function(p) {
  a <- wa |> filter(district == p[1]) |> st_geometry()
  b <- wa |> filter(district == p[2]) |> st_geometry()
  area <- as.numeric(st_area(st_intersection(a, b)))
  tibble(d1 = p[1], d2 = p[2], area_sqm = if (length(area) == 0) 0 else area)
}) |>
  filter(area_sqm > 0) |>
  arrange(desc(area_sqm))
overlaps
```

This reproduces exactly the four overlaps reported in the issue, and finds
no others anywhere else in the file -- confirming District 4 is indeed the
only district with a topology problem here.

## Determining the fix

Plotting one of the overlaps (District 2 vs. District 4, the largest) shows
the overlap isn't a small localized patch, but a thin ribbon running along
almost the *entire* shared boundary -- classic evidence that the two
districts' shared edge was digitized twice, as two slightly different
lines, rather than sharing a single common boundary:

```{r}
#| fig-width: 6
#| fig-height: 8
d2 <- wa |> filter(district == 2) |> st_geometry()
d4 <- wa |> filter(district == 4) |> st_geometry()
overlap_24 <- st_intersection(d2, d4)

ggplot() +
  geom_sf(data = d2, fill = NA, color = "red") +
  geom_sf(data = d4, fill = NA, color = "blue") +
  geom_sf(data = overlap_24, fill = "magenta", color = "magenta") +
  coord_sf(xlim = st_bbox(st_buffer(overlap_24, 0.05))[c("xmin", "xmax")],
           ylim = st_bbox(st_buffer(overlap_24, 0.05))[c("ymin", "ymax")]) +
  labs(title = "District 2 (red) / District 4 (blue), overlap in magenta") +
  theme_minimal()
```

Since District 4 is the only common participant, and none of its four
neighbors overlap *each other*, District 4's neighbors can be trusted as
the authoritative edge here -- consistent with the issue's own diagnosis
that District 4's polygon is the source of the bad edges. Trimming
District 4 back to exclude whatever its neighbors already claim (a plain
`st_difference`) resolves every overlap, but it's not the whole story: it
only removes territory District 4 wrongly claimed, so any spot where
District 4's original line instead fell *short* of a neighbor's true edge
is left as a thin unclaimed gap -- the other side of the same
double-digitized-edge problem.

Closing those gaps by *estimating* how far to extend District 4 (a buffer
distance, a ribbon along a neighbor's line within some tolerance) turned
out to be the wrong tool for the job: every tolerance that was wide enough
to bridge one gap either missed another, narrower one nearby, or -- when
widened further to be safe -- started reaching into unrelated complexity
(merging separate islands together, or sweeping in hundreds of square
kilometers of area that was never actually part of this problem). The
right tool is exact, not approximate: a district boundary vector dataset
like this one should tile the plane with no gaps between adjacent
districts, so *any* point inside the combined footprint of all eight
Washington districts that isn't currently claimed by any of them is
either open water (not assigned to any district anywhere in this file) or
a bug. Computing that set difference directly -- rather than guessing a
distance -- finds every gap adjacent to District 4 in one pass, with nothing
left to tune:

## Fix

```{r}
#| message: false
d1 <- wa |> filter(district == 1) |> st_geometry()
d3 <- wa |> filter(district == 3) |> st_geometry()
d5 <- wa |> filter(district == 5) |> st_geometry()
d6 <- wa |> filter(district == 6) |> st_geometry()
d7 <- wa |> filter(district == 7) |> st_geometry()
d8 <- wa |> filter(district == 8) |> st_geometry()

WA_ALBERS <- 5070 # NAD83 Conus Albers, meters
orig_crs <- st_crs(wa)

to_albers <- function(x) st_transform(x, WA_ALBERS)
d2p <- to_albers(d2); d3p <- to_albers(d3); d4p <- to_albers(d4)
d5p <- to_albers(d5); d8p <- to_albers(d8)
neighbors_p <- st_union(c(d2p, d3p, d5p, d8p)) |> st_make_valid()

# Every point inside 40km of District 4 that isn't currently claimed by any
# of the state's 8 districts. 40km is just a generous search radius (the
# fix below only keeps pieces that actually touch District 4, so a wider
# radius can't over-reach); doubling it to 80km finds nothing new.
all_claimed_p <- st_union(do.call(c, lapply(
  list(d1, d2, d3, d4, d5, d6, d7, d8), to_albers
))) |> st_make_valid()
search_zone <- st_buffer(d4p, 40000)
unclaimed <- st_difference(search_zone, all_claimed_p) |> st_make_valid()

# Real gaps here run from ~400,000 to ~16,000,000 sqm; the Washington/Oregon
# and Washington/Canada state lines also fall within 40km of District 4 in
# places and produce two enormous "unclaimed" pieces (everything outside
# the state) that are an artifact of the search radius, not a bug -- an
# upper bound well below either of those (~2.5 billion sqm) drops them,
# while a lower bound of 1,000 sqm drops reprojection-noise-scale slivers.
# Keeping only pieces that actually touch District 4 additionally excludes
# anything on the far side of a neighbor that isn't really adjacent here.
parts <- st_cast(unclaimed, "MULTIPOLYGON", warn = FALSE) |> st_cast("POLYGON", warn = FALSE)
areas <- as.numeric(st_area(parts))
touches_d4 <- as.numeric(st_distance(parts, d4p)) < 1
real_gaps <- parts[areas > 1000 & areas < 1e8 & touches_d4]
length(real_gaps)
sum(as.numeric(st_area(real_gaps)))

d4_trimmed <- st_difference(d4p, neighbors_p) |> st_make_valid()
d4_filled <- st_union(d4_trimmed, st_union(real_gaps)) |> st_make_valid()
new_d4_geom <- st_difference(d4_filled, neighbors_p) |> st_make_valid() # safety re-trim

# As in the MI 15/16 patch, this kind of union/difference chain can leave
# tiny spurious interior holes that whole-part filtering can't catch since
# they're inside a part that's otherwise kept. District 4 has no real
# interior exclusion at this scale, so drop any hole outright.
strip_holes <- function(x, crs) {
  st_sfc(lapply(st_geometry(x), function(g) {
    parts <- if (inherits(g, "MULTIPOLYGON")) g else list(g)
    st_multipolygon(lapply(parts, function(p) list(p[[1]])))
  }), crs = crs) |> st_make_valid()
}
new_d4_geom <- strip_holes(st_sfc(new_d4_geom, crs = WA_ALBERS), WA_ALBERS)
new_d4_geom <- st_transform(new_d4_geom, orig_crs)

new_d4 <- wa |> filter(district == 4) |> st_set_geometry(new_d4_geom)
```

## Validation

The new District 4 is a valid MultiPolygon with the same six parts as
before (five untouched islands plus the trimmed/extended mainland), no
interior holes, and no overlap with any neighbor:

```{r}
st_is_valid(new_d4)

count_holes <- function(x) {
  sum(sapply(st_geometry(x), function(g) {
    parts <- if (inherits(g, "MULTIPOLYGON")) g else list(g)
    sum(sapply(parts, function(p) length(p) - 1))
  }))
}
count_holes(new_d4)

length(st_cast(st_geometry(new_d4), "MULTIPOLYGON", warn = FALSE)[[1]])

sapply(c(2, 3, 5, 8), function(d) {
  as.numeric(st_area(st_intersection(st_geometry(new_d4), wa |> filter(district == d) |> st_geometry())))
})
```

### Checking for any remaining gaps

Two weaker checks were tried before landing on the exact method above, and
both turned out to be unreliable enough to actually ship: a narrow-corridor
buffer check (only catches gaps narrower than the corridor -- silently
missed everything found and fixed above) and a per-boundary-point distance
sample (an improvement, but still capped by how densely points are
sampled and which stretches of boundary get checked). The direct,
exhaustive version doesn't have either failure mode: re-run the identical
set-difference computation against the *corrected* District 4, and by
construction anything it finds is a real remaining gap:

```{r}
new_d4_p <- to_albers(st_geometry(new_d4))
all_claimed_final <- st_union(c(
  do.call(c, lapply(list(d1, d3, d5, d6, d7, d8), to_albers)), d2p, new_d4_p
)) |> st_make_valid()

remaining_unclaimed <- st_difference(st_buffer(new_d4_p, 40000), all_claimed_final) |>
  st_make_valid()
remaining_parts <- st_cast(remaining_unclaimed, "MULTIPOLYGON", warn = FALSE) |> st_cast("POLYGON", warn = FALSE)
remaining_areas <- as.numeric(st_area(remaining_parts))
remaining_touches <- as.numeric(st_distance(remaining_parts, new_d4_p)) < 1
sum(remaining_areas > 1000 & remaining_areas < 1e8 & remaining_touches)
```

Zero -- every real gap adjacent to the corrected District 4 is gone.

Area change is modest (net of overlap removed and gap filled) and every
other district (including 2, 3, 5, and 8 themselves) is completely
untouched:

```{r}
as.numeric(st_area(d4)) - as.numeric(st_area(new_d4))

wa_other <- wa |> filter(district != 4)
new_wa <- bind_rows(wa_other, new_d4) |> arrange(district)

identical(
  wa |> filter(district != 4) |> st_geometry(),
  new_wa |> filter(district != 4) |> st_geometry()
)
```

Re-plotting the same District 2/4 boundary zoom from above, but with the
corrected District 4, shows the overlap ribbon is gone -- the blue line now
runs cleanly alongside the red one instead of straddling it:

```{r}
#| fig-width: 6
#| fig-height: 8
new_overlap_24 <- st_intersection(d2, st_geometry(new_d4))

ggplot() +
  geom_sf(data = d2, fill = NA, color = "red") +
  geom_sf(data = st_geometry(new_d4), fill = NA, color = "blue") +
  coord_sf(xlim = st_bbox(st_buffer(overlap_24, 0.05))[c("xmin", "xmax")],
           ylim = st_bbox(st_buffer(overlap_24, 0.05))[c("ymin", "ymax")]) +
  labs(title = "Corrected: District 2 (red) / District 4 (blue), no more overlap") +
  theme_minimal()
```

And the full corrected map, all eight districts:

```{r}
#| fig-width: 8
#| fig-height: 8
new_wa |>
  ggplot() +
  geom_sf(aes(fill = as.character(district))) +
  labs(title = "Washington, 99th-102nd Congress (corrected)", fill = "District") +
  theme_minimal()
```

## Visual validation against a real basemap

A closer look at the reported overlaps raised a natural follow-up: are
these overlaps/gaps small relative to real terrain features (roads,
rivers, lakes) the district lines should be following, or does the trimmed
edge look arbitrary? Plotting the corrected District 4 directly on an OSM
basemap answers that.

Along the Columbia River (District 3/4), the corrected boundary hugs the
river channel closely, including its small braids and islands -- exactly
matching the repo owner's original read that these boundaries follow real
water features:

```{r}
#| message: false
#| results: hide
#| fig-width: 9
#| fig-height: 7
d3_4326 <- st_transform(d3, 4326)
new_d4_4326 <- st_transform(st_geometry(new_d4), 4326)

bb_river <- st_bbox(c(xmin = -122.5, xmax = -121.7, ymin = 45.55, ymax = 45.8), crs = 4326)
ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  geom_sf(data = d3_4326, fill = NA, color = "red", linewidth = 1) +
  geom_sf(data = new_d4_4326, fill = NA, color = "blue", linewidth = 1) +
  coord_sf(xlim = bb_river[c("xmin", "xmax")], ylim = bb_river[c("ymin", "ymax")], crs = 4326) +
  labs(title = "District 3 (red) / corrected District 4 (blue) along the Columbia River") +
  theme_minimal()
```

Near the District 2/4 boundary in the remote North Cascades (around the
Chickamin Glacier, well above any road), the corrected line now sits
exactly on top of District 2's -- close enough that only blue is visible,
drawn last -- closing what was actually four separate small gaps here
(totaling ~4,600,000 sqm) that a 150m-tolerance ribbon fix had missed
entirely, since real-world offsets in this stretch turned out to run
wider than that in places. There's no road/river to follow here, so the
line most likely follows a section-line or ridge, but critically the
exact fix doesn't introduce any spurious zigzags getting there:

```{r}
#| message: false
#| results: hide
#| fig-width: 9
#| fig-height: 9
d2_4326 <- st_transform(d2, 4326)

bb_glacier <- st_bbox(c(xmin = -121.3, xmax = -120.9, ymin = 48.25, ymax = 48.45), crs = 4326)
ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  geom_sf(data = d2_4326, fill = NA, color = "red", linewidth = 1) +
  geom_sf(data = new_d4_4326, fill = NA, color = "blue", linewidth = 1) +
  coord_sf(xlim = bb_glacier[c("xmin", "xmax")], ylim = bb_glacier[c("ymin", "ymax")], crs = 4326) +
  labs(title = "District 2 (red) / corrected District 4 (blue) near Chickamin Glacier") +
  theme_minimal()
```

The largest single piece the exact fix found is a ~16,000,000 sqm wedge
against District 2 around the Suiattle/White Chuck/Honeycomb glaciers,
where District 4's original line cut a crude straight diagonal across
terrain District 2's line follows in real, glacier-by-glacier detail.
Plotting both versions confirms it: the original line (gray dashed)
ignores the glacier outlines entirely, while the corrected one (blue) now
sits right on top of District 2's detailed line:

```{r}
#| message: false
#| results: hide
#| fig-width: 9
#| fig-height: 7
d4_orig_4326 <- st_transform(d4, 4326)

bb_suiattle <- st_bbox(c(xmin = -121.28, xmax = -120.98, ymin = 47.93, ymax = 48.13), crs = 4326)
ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  geom_sf(data = d4_orig_4326, fill = NA, color = "gray40", linewidth = 0.8, linetype = "dashed") +
  geom_sf(data = d2_4326, fill = NA, color = "red", linewidth = 1) +
  geom_sf(data = new_d4_4326, fill = NA, color = "blue", linewidth = 1) +
  coord_sf(xlim = bb_suiattle[c("xmin", "xmax")], ylim = bb_suiattle[c("ymin", "ymax")], crs = 4326) +
  labs(title = "District 2 (red) / corrected District 4 (blue) / original District 4 (gray) near the Suiattle glaciers") +
  theme_minimal()
```

A second wedge, further south around 47.5N, looked just as large at first
glance -- almost all of it turned out to be a false alarm: that area is
already District 8's territory (just not plotted in the District 2/4
comparison above), so there was nothing to fix there. But not quite all of
it: the exact method also caught one small genuine sliver right at the
District 2/4/8 corner (~1,900,000 sqm) that the eyeballed false alarm would
have hidden completely -- exactly the kind of mistake that motivated
computing the exact unclaimed area instead of estimating it.

(Two earlier, weaker approaches were tried and dropped before landing on
the exact method above. Buffering District 4's *whole* polygon outward and
eroding it back entangled unrelated lake/glacier shoreline detail and
produced spurious zigzags right in this area. A per-neighbor ribbon with a
hand-picked tolerance did better, but every tolerance wide enough to
bridge one gap either missed a narrower one nearby or started reaching
into unrelated territory elsewhere -- which is what motivated computing
the unclaimed area exactly, with nothing left to tune.)

## The District 6 / District 8 report

A later comment on the issue reported District 6 overlapping District 8,
with the overlapping piece of District 6 non-contiguous from the rest of
District 6. Checking directly against the current file finds no overlap
between District 6 and District 8 at all:

```{r}
as.numeric(st_area(st_intersection(d6, d8)))
```

District 6 does have many disconnected parts (it's a Puget Sound district
with many islands, so some fragmentation is expected), but none of them
overlap District 8. This isn't reproducible against the current file, so
it's left out of this patch -- it may have already been resolved by
whatever produced the current file, or the report may have been against a
stale copy of the data. If it recurs, it likely needs its own issue/patch
rather than folding it in here.

## Write out the corrected file

```{r}
unlink(GEOJSON_FILE)
st_write(new_wa, GEOJSON_FILE, append = FALSE)
```
