---
title: "Fix Washington CD 4 Overlaps From Issue #18"
format: html
editor: source
---

## Overview

[Issue #18](https://github.com/JeffreyBLewis/congressional-district-boundaries/issues/18)
reports that in `Washington_099_to_102.geojson` (99th-102nd Congress,
1985-1993), District 4's polygon overlaps four neighboring districts by
small but non-zero amounts:

| Pair | Intersection (deg²) | ~Sq mi |
|---|---|---|
| WA-02 ∩ WA-04 | 0.000846 | ~2.7 |
| WA-03 ∩ WA-04 | 0.000144 | ~0.5 |
| WA-04 ∩ WA-05 | 0.000003 | ~0.01 |
| WA-04 ∩ WA-08 | 0.000014 | ~0.05 |

The reporter noted that District 4 is the common participant in every
overlap, suggesting District 4's own polygon is the source of the bad
edges rather than each neighbor independently. A follow-up comment from
the repo owner confirmed the overlaps are small relative to district size
and fall along water bodies (mainly the Columbia River), and flagged this
as part of a broader topology-cleanup effort. A later comment from a third
party additionally reported District 6 overlapping District 8 with a
non-contiguous piece of District 6; that's checked and addressed separately
below since it turns out not to be reproducible against the current file.

```{r}
#| message: false
library(tidyverse)
library(sf)
library(ggspatial)
sf_use_s2(FALSE)
```

## Confirming the bug

```{r}
GEOJSON_FILE <- "../../GeoJson/Washington_099_to_102.geojson"
wa <- st_read(GEOJSON_FILE, quiet = TRUE)

all_pairs <- combn(sort(unique(wa$district)), 2, simplify = FALSE)
overlaps <- map_dfr(all_pairs, function(p) {
  a <- wa |> filter(district == p[1]) |> st_geometry()
  b <- wa |> filter(district == p[2]) |> st_geometry()
  area <- as.numeric(st_area(st_intersection(a, b)))
  tibble(d1 = p[1], d2 = p[2], area_sqm = if (length(area) == 0) 0 else area)
}) |>
  filter(area_sqm > 0) |>
  arrange(desc(area_sqm))
overlaps
```

This reproduces exactly the four overlaps reported in the issue, and finds
no others anywhere else in the file -- confirming District 4 is indeed the
only district with a topology problem here.

## Determining the fix

Plotting one of the overlaps (District 2 vs. District 4, the largest) shows
the overlap isn't a small localized patch, but a thin ribbon running along
almost the *entire* shared boundary -- classic evidence that the two
districts' shared edge was digitized twice, as two slightly different
lines, rather than sharing a single common boundary:

```{r}
#| fig-width: 6
#| fig-height: 8
d2 <- wa |> filter(district == 2) |> st_geometry()
d4 <- wa |> filter(district == 4) |> st_geometry()
overlap_24 <- st_intersection(d2, d4)

ggplot() +
  geom_sf(data = d2, fill = NA, color = "red") +
  geom_sf(data = d4, fill = NA, color = "blue") +
  geom_sf(data = overlap_24, fill = "magenta", color = "magenta") +
  coord_sf(xlim = st_bbox(st_buffer(overlap_24, 0.05))[c("xmin", "xmax")],
           ylim = st_bbox(st_buffer(overlap_24, 0.05))[c("ymin", "ymax")]) +
  labs(title = "District 2 (red) / District 4 (blue), overlap in magenta") +
  theme_minimal()
```

Since District 4 is the only common participant, and none of its four
neighbors overlap *each other*, District 4's neighbors can be trusted as
the authoritative edge here -- consistent with the issue's own diagnosis
that District 4's polygon is the source of the bad edges. Trimming
District 4 back to exclude whatever its neighbors already claim (a plain
`st_difference`) resolves every overlap, but it's not the whole story: it
only removes territory District 4 wrongly claimed, so any spot where
District 4's original line instead fell *short* of a neighbor's true edge
is left as a thin unclaimed gap -- the other side of the same
double-digitized-edge problem. Since a neighbor's own boundary is trusted
here, the more complete fix is to snap District 4's edge directly onto
each neighbor's actual boundary line wherever they run close together,
rather than just subtracting: build a thin ribbon that traces each
neighbor's true boundary near District 4, add it to District 4, then
subtract the neighbors themselves -- so the result is bounded exactly by
the neighbor's real edge, closing gaps and overlaps alike, without
resorting to a blunt buffer-the-whole-polygon approach (tried first, and
dropped -- see the basemap check below for why).

## Fix

```{r}
#| message: false
d3 <- wa |> filter(district == 3) |> st_geometry()
d5 <- wa |> filter(district == 5) |> st_geometry()
d8 <- wa |> filter(district == 8) |> st_geometry()

WA_ALBERS <- 5070 # NAD83 Conus Albers, meters
orig_crs <- st_crs(wa)

to_albers <- function(x) st_transform(x, WA_ALBERS)
d2p <- to_albers(d2); d3p <- to_albers(d3); d4p <- to_albers(d4)
d5p <- to_albers(d5); d8p <- to_albers(d8)
neighbors_p <- st_union(c(d2p, d3p, d5p, d8p)) |> st_make_valid()

# For each neighbor, a thin ribbon along *that neighbor's own* boundary,
# restricted to the stretch actually near District 4 (so it can't reach
# into a neighbor's boundary with some other, unrelated district). 150m is
# comfortably wider than the gaps/overlaps found above (worst case ~40m
# average width over the D2 boundary) without being so wide it starts
# picking up unrelated detail.
RIBBON_TOL <- 150
ribbons <- lapply(list(d2p, d3p, d5p, d8p), function(nb) {
  near_d4 <- st_intersection(st_boundary(nb), st_buffer(d4p, RIBBON_TOL))
  st_buffer(near_d4, RIBBON_TOL) |> st_make_valid()
})
d4_filled <- st_union(d4p, st_union(do.call(c, ribbons))) |> st_make_valid()
new_d4_geom <- st_difference(d4_filled, neighbors_p) |> st_make_valid()

# As in the MI 15/16 patch, this kind of union/difference chain can leave
# tiny spurious interior holes that whole-part filtering can't catch since
# they're inside a part that's otherwise kept. District 4 has no real
# interior exclusion at this scale, so drop any hole outright.
strip_holes <- function(x, crs) {
  st_sfc(lapply(st_geometry(x), function(g) {
    parts <- if (inherits(g, "MULTIPOLYGON")) g else list(g)
    st_multipolygon(lapply(parts, function(p) list(p[[1]])))
  }), crs = crs) |> st_make_valid()
}
new_d4_geom <- strip_holes(st_sfc(new_d4_geom, crs = WA_ALBERS), WA_ALBERS)
new_d4_geom <- st_transform(new_d4_geom, orig_crs)

new_d4 <- wa |> filter(district == 4) |> st_set_geometry(new_d4_geom)
```

## Validation

The new District 4 is a valid MultiPolygon with the same six parts as
before (five untouched islands plus the trimmed/extended mainland), no
interior holes, and no overlap with any neighbor:

```{r}
st_is_valid(new_d4)

count_holes <- function(x) {
  sum(sapply(st_geometry(x), function(g) {
    parts <- if (inherits(g, "MULTIPOLYGON")) g else list(g)
    sum(sapply(parts, function(p) length(p) - 1))
  }))
}
count_holes(new_d4)

length(st_cast(st_geometry(new_d4), "MULTIPOLYGON", warn = FALSE)[[1]])

sapply(c(2, 3, 5, 8), function(d) {
  as.numeric(st_area(st_intersection(st_geometry(new_d4), wa |> filter(district == d) |> st_geometry())))
})
```

And, unlike a plain trim, the gaps along these same boundaries are now
closed too -- checking within a 5m margin (tight enough that only a
genuine unclosed gap would show up, not floating-point/reprojection noise)
finds none of any consequence, versus tens to hundreds of thousands of
square meters before fixing:

```{r}
gap_area <- function(a, b, tol_m) {
  a_p <- st_transform(a, WA_ALBERS); b_p <- st_transform(b, WA_ALBERS)
  zone <- st_intersection(st_buffer(a_p, tol_m), st_buffer(b_p, tol_m))
  as.numeric(st_area(st_difference(st_difference(zone, a_p), b_p)))
}
sapply(c(2, 3, 5, 8), function(d) {
  gap_area(st_geometry(new_d4), wa |> filter(district == d) |> st_geometry(), 5)
})
```

Area change is modest (net of overlap removed and gap filled) and every
other district (including 2, 3, 5, and 8 themselves) is completely
untouched:

```{r}
as.numeric(st_area(d4)) - as.numeric(st_area(new_d4))

wa_other <- wa |> filter(district != 4)
new_wa <- bind_rows(wa_other, new_d4) |> arrange(district)

identical(
  wa |> filter(district != 4) |> st_geometry(),
  new_wa |> filter(district != 4) |> st_geometry()
)
```

Re-plotting the same District 2/4 boundary zoom from above, but with the
corrected District 4, shows the overlap ribbon is gone -- the blue line now
runs cleanly alongside the red one instead of straddling it:

```{r}
#| fig-width: 6
#| fig-height: 8
new_overlap_24 <- st_intersection(d2, st_geometry(new_d4))

ggplot() +
  geom_sf(data = d2, fill = NA, color = "red") +
  geom_sf(data = st_geometry(new_d4), fill = NA, color = "blue") +
  coord_sf(xlim = st_bbox(st_buffer(overlap_24, 0.05))[c("xmin", "xmax")],
           ylim = st_bbox(st_buffer(overlap_24, 0.05))[c("ymin", "ymax")]) +
  labs(title = "Corrected: District 2 (red) / District 4 (blue), no more overlap") +
  theme_minimal()
```

And the full corrected map, all eight districts:

```{r}
#| fig-width: 8
#| fig-height: 8
new_wa |>
  ggplot() +
  geom_sf(aes(fill = as.character(district))) +
  labs(title = "Washington, 99th-102nd Congress (corrected)", fill = "District") +
  theme_minimal()
```

## Visual validation against a real basemap

A closer look at the reported overlaps raised a natural follow-up: are
these overlaps/gaps small relative to real terrain features (roads,
rivers, lakes) the district lines should be following, or does the trimmed
edge look arbitrary? Plotting the corrected District 4 directly on an OSM
basemap answers that.

Along the Columbia River (District 3/4), the corrected boundary hugs the
river channel closely, including its small braids and islands -- exactly
matching the repo owner's original read that these boundaries follow real
water features:

```{r}
#| message: false
#| results: hide
#| fig-width: 9
#| fig-height: 7
d3_4326 <- st_transform(d3, 4326)
new_d4_4326 <- st_transform(st_geometry(new_d4), 4326)

bb_river <- st_bbox(c(xmin = -122.5, xmax = -121.7, ymin = 45.55, ymax = 45.8), crs = 4326)
ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  geom_sf(data = d3_4326, fill = NA, color = "red", linewidth = 1) +
  geom_sf(data = new_d4_4326, fill = NA, color = "blue", linewidth = 1) +
  coord_sf(xlim = bb_river[c("xmin", "xmax")], ylim = bb_river[c("ymin", "ymax")], crs = 4326) +
  labs(title = "District 3 (red) / corrected District 4 (blue) along the Columbia River") +
  theme_minimal()
```

Near the District 2/4 boundary in the remote North Cascades (around the
Chickamin Glacier, well above any road), the corrected line still runs
cleanly alongside District 2 with no jagged artifacts -- there's no
road/river to follow here, so the line most likely follows a section-line
or ridge, but critically the fix doesn't introduce any spurious zigzags:

```{r}
#| message: false
#| results: hide
#| fig-width: 9
#| fig-height: 9
d2_4326 <- st_transform(d2, 4326)

bb_glacier <- st_bbox(c(xmin = -121.3, xmax = -120.9, ymin = 48.25, ymax = 48.45), crs = 4326)
ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  geom_sf(data = d2_4326, fill = NA, color = "red", linewidth = 1) +
  geom_sf(data = new_d4_4326, fill = NA, color = "blue", linewidth = 1) +
  coord_sf(xlim = bb_glacier[c("xmin", "xmax")], ylim = bb_glacier[c("ymin", "ymax")], crs = 4326) +
  labs(title = "District 2 (red) / corrected District 4 (blue) near Chickamin Glacier") +
  theme_minimal()
```

(An earlier version of this fix closed gaps by buffering District 4's
*whole* polygon outward and eroding it back -- a blunt instrument that,
right in this glacier area, produced exactly the kind of spurious zigzag
this section checks for, by entangling nearby lake/glacier shoreline
detail it had no business touching. Snapping a thin ribbon directly onto
each neighbor's own boundary line instead -- the version fixed above --
avoids that: it only ever touches the narrow strip actually adjacent to
the true edge, so it can't reach into unrelated complexity a few hundred
meters away.)

## The District 6 / District 8 report

A later comment on the issue reported District 6 overlapping District 8,
with the overlapping piece of District 6 non-contiguous from the rest of
District 6. Checking directly against the current file finds no overlap
between District 6 and District 8 at all:

```{r}
d6 <- wa |> filter(district == 6) |> st_geometry()
as.numeric(st_area(st_intersection(d6, d8)))
```

District 6 does have many disconnected parts (it's a Puget Sound district
with many islands, so some fragmentation is expected), but none of them
overlap District 8. This isn't reproducible against the current file, so
it's left out of this patch -- it may have already been resolved by
whatever produced the current file, or the report may have been against a
stale copy of the data. If it recurs, it likely needs its own issue/patch
rather than folding it in here.

## Write out the corrected file

```{r}
unlink(GEOJSON_FILE)
st_write(new_wa, GEOJSON_FILE, append = FALSE)
```
