---
title: "Fix Michigan CD 15/16 Boundary From Issue #24"
format: html
editor: source
---

## Overview

Benjamin Dasilva reported in [Issue #24](https://github.com/JeffreyBLewis/congressional-district-boundaries/issues/24)
that in `Michigan_089_to_092.geojson` (89th-92nd Congress, 1965-1972), the
suburban ("Downriver") portion of District 16 is incorrectly included in
District 15, and that the on-file District 16 shape only covers "the Detroit
portion of the district."

```{r}
#| message: false
library(tidyverse)
library(sf)
library(tigris)
library(readxl)
options(tigris_use_cache = TRUE)
sf_use_s2(FALSE)
```

## Confirming the bug

```{r}
GEOJSON_FILE <- "../../GeoJson/Michigan_089_to_092.geojson"
mi <- st_read(GEOJSON_FILE, quiet = TRUE)
d15 <- mi |> filter(district == 15)
d16 <- mi |> filter(district == 16)
```

On file, `district == 16` is tiny (`{r} as.numeric(st_area(d16))` sq. m,
vs. `{r} as.numeric(st_area(d15))` for `district == 15`). Comparing it to
current Wayne County, MI civil boundaries (which for these long-incorporated
cities are essentially unchanged since the 1960s) shows what it actually is:

```{r}
wayne_cousub <- county_subdivisions(state = "MI", county = "Wayne", cb = TRUE, year = 2020) |>
  st_transform(st_crs(mi))

detroit <- wayne_cousub |> filter(NAME == "Detroit")
as.numeric(st_area(st_intersection(st_geometry(d16), st_geometry(detroit)))) /
  as.numeric(st_area(d16))
```

Essentially all (>99.9%) of the on-file District 16 shape sits inside the
City of Detroit -- specifically, its irregular southwestern ("Delray")
panhandle that pokes down between Melvindale, Dearborn, and River Rouge
along the Rouge River. That panhandle is geographically cut off from the
rest of Detroit by the surrounding suburbs, so it plausibly really was
drawn into the surrounding 16th district. But essentially none of the
Downriver suburbs that should make up the bulk of District 16 are present.

## Determining the correct boundary

The Census Bureau's scanned map of Wayne County for the 89th Congress
(attached to the issue) shows District 16's approximate outline, but hand
tracing which side of a hand-drawn, 55-year-old scanned line a given city
falls on is error-prone -- an initial pass at this patch misread the line
around Melvindale/Allen Park/Lincoln Park from the scan alone and got it
wrong. The same UCLA project that produced this repository's source data
also published a per-state, per-era notes file documenting exactly which
counties/townships/cities were assigned to each district, citing the
actual 1964 Michigan apportionment law (Act 282) as its source. That's a
much more reliable, unambiguous reference than reading a scan, so it's
used here as ground truth:

```{r}
notes_url <- "https://cdmaps.polisci.ucla.edu/doc/89-92cc_Michigan_Notes.xlsx"
notes_file <- tempfile(fileext = ".xlsx")
download.file(notes_url, notes_file, quiet = TRUE, mode = "wb")

raw <- read_excel(notes_file, sheet = "Sheet1", col_names = FALSE)
notes <- raw |>
  filter(row_number() > which(raw[[1]] == "State")) |>
  set_names(c("state", "district", "county", "x4", "x5", "x6", "progress", "note")) |>
  filter(district %in% c("15", "16"), county == "Wayne") |>
  select(district, note)

notes |> pull(note) |> walk(\(x) {cat(strwrap(x, width = 90), sep = "\n"); cat("\n\n")})
```

District 15's Wayne County territory is: townships of Canton, Dearborn,
Huron, Nankin, Romulus, Sumpter, Taylor, Van Buren; cities of Allen Park,
Belleville, Dearborn Heights, Garden City, Lincoln Park, Melvindale, Wayne.

District 16's Wayne County territory is: townships of Brownstown and Grosse
Ile; cities of Dearborn, Ecorse, Gibraltar, River Rouge, Riverview,
Southgate, Trenton, Wyandotte; plus the part of the City of Detroit
described as the panhandle already correctly on file (confirmed above).

So Melvindale, Allen Park, and Lincoln Park -- despite sitting geographically
between Dearborn and the rest of Downriver -- are District 15, forming a
peninsula that pokes into what's otherwise District 16 territory. Woodhaven,
Flat Rock, and Rockwood aren't named explicitly (they incorporated out of
Brownstown Township in 1965-68, shortly after the 1964 law was written), but
nothing in the notes or the map shows any district line cutting them off
from the rest of the Downriver cluster, so they're treated as inheriting
Brownstown Township's District 16 assignment.

```{r}
d16_places <- c("Dearborn", "Ecorse", "River Rouge", "Southgate", "Wyandotte",
                "Riverview", "Trenton", "Gibraltar", "Woodhaven", "Flat Rock",
                "Rockwood", "Brownstown", "Grosse Ile")
target <- wayne_cousub |>
  filter(NAME %in% d16_places) |>
  st_union() |>
  st_make_valid()

tibble(
  district = c(15, 16),
  area_sqm = c(as.numeric(st_area(d15)), as.numeric(st_area(d16))),
  frac_of_target_cities = c(
    as.numeric(st_area(st_intersection(st_geometry(d15), target))) / as.numeric(st_area(target)),
    as.numeric(st_area(st_intersection(st_geometry(d16), target))) / as.numeric(st_area(target))
  )
)
```

96% of the cities/townships that should be District 16 are instead part of
District 15 on file, confirming the bug.

Checking the neighboring era files (`Michigan_093_to_097.geojson`,
`Michigan_098_to_102.geojson`, `Michigan_103_to_107.geojson`) shows their
`district == 15`/`16` boundaries were redrawn at each subsequent
redistricting (Michigan's district lines here were not static), so they
aren't used to infer the 89-92 boundary -- the notes file is the only
source used for that.

## Fix

Move the correct District 16 cities/townships (clipped to the existing
District 15/16 combined area, so we don't take territory from neighboring
districts 1 or 17) from District 15 to District 16, keeping the existing
(correct) Detroit panhandle piece as part of District 16 as well.

Doing this with a naive `st_intersection`/`st_difference` leaves visible
artifacts, for two separate reasons found while checking the result against
the Census map:

  - `wayne_cousub`'s coastline (modern TIGER data) doesn't exactly line up
    with the 1970 Census shapefile's coastline along the Detroit River, and
    cutting along it directly leaves a scatter of sliver artifacts on the
    shore.
  - The original (pre-fix, untouched) `district == 15` shape already
    contains a spurious ~2km-long, ~30m-wide spike near Ecorse -- a
    pre-existing digitizing glitch in the 1970 source data, unrelated to
    this bug, but this patch is already rebuilding this exact area so it's
    addressed here too rather than left in place.

Both are fixed the same way: work in a projected CRS (Michigan State Plane
South, so distances are in meters and the cleanup thresholds below mean
what they say), and after each boundary operation, keep only real polygon
area and discard anything under a small size threshold -- both stray
degenerate slivers and, in the one case above, the pre-existing spike. The
result is transformed back to the file's own CRS for writing; the
projection is just a computational tool for this step, not something that
ends up in the data.

```{r}
MI_SOUTH <- 6498 # NAD83(2011) / Michigan South, meters
orig_crs <- st_crs(mi)

d15_p <- st_transform(d15, MI_SOUTH)
d16_p <- st_transform(d16, MI_SOUTH)
target_p <- st_transform(target, MI_SOUTH)

# Strip the pre-existing spike from the combined D15/D16 outline once, up
# front, via a morphological "opening" (erode then dilate by the same
# radius): anything narrower than 2x the radius is removed, wider features
# are left intact. 25m safely clears the ~30m-wide spike while being far
# smaller than any real neighborhood/parcel here.
region_raw <- st_union(st_geometry(d15_p), st_geometry(d16_p)) |> st_make_valid()
region <- st_buffer(st_buffer(region_raw, -25), 25) |> st_make_valid()
d15_p <- st_intersection(st_geometry(d15_p), region) |> st_make_valid()
d16_p <- st_intersection(st_geometry(d16_p), region) |> st_make_valid()

target_snapped <- st_snap(target_p, st_boundary(region), tolerance = 75) |> st_make_valid()
target_clip <- st_intersection(target_snapped, region) |> st_make_valid()

new_d16_geom <- st_union(d16_p, target_clip) |> st_make_valid()
new_d15_geom <- st_difference(d15_p, target_clip) |> st_make_valid()

# Keep only real polygon area: st_intersection/st_difference between two
# slightly-mismatched coastlines can return a GEOMETRYCOLLECTION mixing in
# degenerate LINESTRING/POINT slivers, and even the polygon parts can
# include small coastline-mismatch fragments -- both would otherwise show
# up as stray marks when plotted. 25,000 sqm sits well below every real
# named place here (the smallest true river island in the final result
# runs ~30,000 sqm) and well above the mismatch slivers (which top out
# around 20,000 sqm and are much less circular -- true islands are
# comparatively round, slivers are thin ribbons).
clean_polys <- function(g, crs, min_area_sqm = 25000) {
  g <- st_sfc(g, crs = crs) |> st_collection_extract("POLYGON") |> st_make_valid()
  parts <- st_cast(g, "MULTIPOLYGON", warn = FALSE) |> st_cast("POLYGON", warn = FALSE)
  parts <- parts[as.numeric(st_area(parts)) >= min_area_sqm]
  st_sfc(st_union(parts) |> st_make_valid(), crs = crs)
}

new_d15 <- st_set_geometry(d15, st_transform(clean_polys(new_d15_geom, MI_SOUTH), orig_crs))
new_d16 <- st_set_geometry(d16, st_transform(clean_polys(new_d16_geom, MI_SOUTH), orig_crs))

# Reprojection itself can reintroduce degenerate slivers, so repeat the
# same polygon-only, small-fragment cleanup once more after transforming
# back (a 1 sqm floor here -- this pass is only mopping up reprojection
# noise, not real coastline mismatch).
st_geometry(new_d15) <- clean_polys(st_geometry(new_d15)[[1]], orig_crs, min_area_sqm = 1)
st_geometry(new_d16) <- clean_polys(st_geometry(new_d16)[[1]], orig_crs, min_area_sqm = 1)
```

## Validation

Both shapes are valid simple-feature polygons (not stray line/point
artifacts), area is conserved (up to the sub-parcel slivers dropped above),
and the two new shapes don't overlap:

```{r}
all(st_is_valid(new_d15), st_is_valid(new_d16))
as.numeric(st_area(region))
as.numeric(st_area(new_d15)) + as.numeric(st_area(new_d16))
as.numeric(st_area(st_intersection(new_d15, new_d16)))
```

District 16 now covers essentially all of its target cities/townships, and
District 15 essentially none of them:

```{r}
tibble(
  district = c(15, 16),
  frac_of_target_cities = c(
    as.numeric(st_area(st_intersection(st_geometry(new_d15), target))) / as.numeric(st_area(target)),
    as.numeric(st_area(st_intersection(st_geometry(new_d16), target))) / as.numeric(st_area(target))
  )
)
```

And each individual city/township lands in the district the notes call for
(cross-checking Melvindale, Allen Park, and Lincoln Park stayed in District
15 rather than following their Downriver neighbors into District 16). A few
of the Downriver cities/townships fall short of 100% in District 16 -- that's
their share of open water (the Detroit River channel), which, like every
other district in this dataset (checked against District 13's Detroit
shoreline), isn't assigned to any district at all, not a bug here:

```{r}
check_places <- c("Melvindale", "Allen Park", "Lincoln Park", "Dearborn",
                   "Ecorse", "Wyandotte", "Grosse Ile", "Rockwood")
map_dfr(check_places, function(nm) {
  row <- wayne_cousub |> filter(NAME == nm)
  cg <- st_geometry(row) |> st_union()
  tibble(
    place = nm,
    frac_water = row$AWATER[1] / (row$ALAND[1] + row$AWATER[1]),
    frac_in_d15 = as.numeric(st_area(st_intersection(cg, st_geometry(new_d15)))) / as.numeric(st_area(cg)),
    frac_in_d16 = as.numeric(st_area(st_intersection(cg, st_geometry(new_d16)))) / as.numeric(st_area(cg))
  )
})
```

## Write out the corrected file

```{r}
out <- mi |>
  filter(!(district %in% c(15, 16))) |>
  bind_rows(new_d15, new_d16) |>
  arrange(district)

unlink(GEOJSON_FILE)
st_write(out, GEOJSON_FILE, append = FALSE)
```

## Visual validation against the Census map

As a quick sanity check, overlay the corrected District 15/16 shapes
directly on top of the Census Bureau's own scan of the 89th Congress Wayne
County map (the one attached to the issue). The scan has no embedded
coordinates, so it first has to be georeferenced: three small,
distinctively-shaped municipalities that are easy to pick out by eye in the
scan -- Plymouth city, Belleville city, and the Highland Park enclave -- and
whose real-world boundaries have not meaningfully changed since the 1960s,
are used as control points (their pixel bounding boxes in the scan, matched
to their `tigris` bounding boxes) to fit a pixel -> map transform.

This fit (and the plot below) is done in the same Michigan State Plane
South projection used for the fix above, rather than raw lon/lat -- that's
the natural projection for an official Michigan county-level map like this
one, and it keeps distances/angles locally undistorted. This is purely for
drawing the comparison figure; it has no bearing on the `Fix` section above
or the coordinates written to the GeoJSON file.

```{r}
#| message: false
map_url <- "https://github.com/user-attachments/assets/a558df0c-39b7-4198-ab3b-647be0f32307"
map_file <- tempfile(fileext = ".png")
download.file(map_url, map_file, quiet = TRUE, mode = "wb")

img <- png::readPNG(map_file)
img_h <- dim(img)[1]
img_w <- dim(img)[2]

# Control points: pixel bounding-box corners (col, row) of Plymouth city,
# Belleville city, and the Highland Park enclave, read off the scan, matched
# to each city's real (NW, SE) bounding-box corner from `wayne_cousub`
# (confirmed via st_bbox() on each city, filtered to NAMELSAD == "... city"
# to avoid the same-named surrounding township), reprojected into Michigan
# State Plane South (meters).
gcp <- tibble(
  place = c("Plymouth NW", "Plymouth SE", "Belleville NW", "Belleville SE",
            "Highland Park NW", "Highland Park SE"),
  px    = c(208, 248, 190, 222, 657, 726),
  py    = c(191, 228, 467, 502, 140, 190),
  lon   = c(-83.48687, -83.44552, -83.49901, -83.47237, -83.12193, -83.07377),
  lat   = c(42.38445, 42.35850, 42.21386, 42.19006, 42.41816, 42.38808)
)
gcp_xy <- gcp |>
  st_as_sf(coords = c("lon", "lat"), crs = orig_crs) |>
  st_transform(MI_SOUTH) |>
  st_coordinates() |>
  as_tibble() |>
  bind_cols(gcp |> select(place, px, py))

x_fit <- lm(X ~ px, data = gcp_xy)
y_fit <- lm(Y ~ py, data = gcp_xy)

max(abs(residuals(x_fit)))
max(abs(residuals(y_fit)))

xmin <- predict(x_fit, tibble(px = 0))
xmax <- predict(x_fit, tibble(px = img_w))
ymax <- predict(y_fit, tibble(py = 0))
ymin <- predict(y_fit, tibble(py = img_h))
```

Residuals are well under 500m at county scale, plenty precise for a visual
check. Now plot the corrected shapes (reprojected to the same CRS) on top
of the georeferenced scan:

```{r}
#| fig-width: 9
#| fig-height: 9
new_d15_p <- st_transform(new_d15, MI_SOUTH)
new_d16_p <- st_transform(new_d16, MI_SOUTH)

ggplot() +
  annotation_raster(img, xmin, xmax, ymin, ymax) +
  geom_sf(data = new_d15_p, fill = NA, color = "red", linewidth = 1) +
  geom_sf(data = new_d16_p, fill = NA, color = "blue", linewidth = 1) +
  coord_sf(crs = MI_SOUTH, xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  labs(title = "Corrected District 15 (red) / District 16 (blue) vs. Census map") +
  theme_minimal()
```

And zoomed in on Allen Park/Melvindale/Lincoln Park, to check that peninsula
specifically against the map:

```{r}
#| fig-width: 8
#| fig-height: 8
zoom_bb <- st_bbox(c(xmin = -83.22, xmax = -83.12, ymin = 42.25, ymax = 42.35), crs = orig_crs) |>
  st_as_sfc() |>
  st_transform(MI_SOUTH) |>
  st_bbox()

ggplot() +
  annotation_raster(img, xmin, xmax, ymin, ymax) +
  geom_sf(data = new_d15_p, fill = NA, color = "red", linewidth = 1.2) +
  geom_sf(data = new_d16_p, fill = NA, color = "blue", linewidth = 1.2) +
  coord_sf(crs = MI_SOUTH, xlim = zoom_bb[c("xmin", "xmax")], ylim = zoom_bb[c("ymin", "ymax")], expand = FALSE) +
  labs(title = "Zoom: Allen Park / Melvindale / Lincoln Park / Ecorse") +
  theme_minimal()
```

The corrected District 16 (blue) traces the Census map's Downriver outline
as clean, closed polygon edges (no stray sliver marks), including the full
extent south to Rockwood/Flat Rock, while District 15 (red) keeps its
Allen Park/Melvindale/Lincoln Park peninsula intact.
