---
title: "Fix Michigan CD 15/16 Boundary From Issue #24"
format: html
editor: source
---

## Overview

Benjamin Dasilva reported in [Issue #24](https://github.com/JeffreyBLewis/congressional-district-boundaries/issues/24)
that in `Michigan_089_to_092.geojson` (89th-92nd Congress, 1965-1972), the
suburban ("Downriver") portion of District 16 is incorrectly included in
District 15, and that the on-file District 16 shape only covers "the Detroit
portion of the district."

```{r}
#| message: false
library(tidyverse)
library(sf)
library(tigris)
options(tigris_use_cache = TRUE)
sf_use_s2(FALSE)
```

## Confirming the bug

```{r}
GEOJSON_FILE <- "../../GeoJson/Michigan_089_to_092.geojson"
mi <- st_read(GEOJSON_FILE, quiet = TRUE)
d15 <- mi |> filter(district == 15)
d16 <- mi |> filter(district == 16)
```

The Census Bureau's own map of Wayne County for the 89th Congress (attached
to the issue) shows District 16 as a compact area covering the city of
Dearborn and the inner Downriver suburbs -- Melvindale, Allen Park, Lincoln
Park, River Rouge, Ecorse, Southgate, and Wyandotte -- with District 15
wrapping around it to the west, north, and south (Canton, Nankin, Garden
City, Dearborn Heights, Inkster, Wayne, Romulus, Taylor, and the outer
Downriver communities further south like Trenton, Woodhaven, and Rockwood).

On file, `district == 16` is tiny (`{r} as.numeric(st_area(d16))` sq. m,
vs. `{r} as.numeric(st_area(d15))` for `district == 15`) and its shape does
not resemble Dearborn/Downriver at all. Comparing it to current Wayne
County, MI civil boundaries (which for these long-incorporated cities are
essentially unchanged since the 1960s) confirms what it actually is:

```{r}
wayne_cousub <- county_subdivisions(state = "MI", county = "Wayne", cb = TRUE, year = 2020) |>
  st_transform(st_crs(mi))

detroit <- wayne_cousub |> filter(NAME == "Detroit")
as.numeric(st_area(st_intersection(st_geometry(d16), st_geometry(detroit)))) /
  as.numeric(st_area(d16))
```

Essentially all (>99.9%) of the on-file District 16 shape sits inside the
City of Detroit -- specifically, its irregular southwestern ("Delray")
panhandle that pokes down between Melvindale, Dearborn, and River Rouge
along the Rouge River. That panhandle is geographically cut off from the
rest of Detroit by the surrounding suburbs, so it is plausible it really
was drawn into the surrounding 16th district. But the suburban cities that
should make up the bulk of District 16 -- Dearborn, Melvindale, Allen Park,
Lincoln Park, River Rouge, Ecorse, Southgate, and Wyandotte -- are missing
from it entirely:

```{r}
d16_cities <- c("Dearborn", "Melvindale", "Allen Park", "Lincoln Park",
                "River Rouge", "Ecorse", "Southgate", "Wyandotte")
target <- wayne_cousub |>
  filter(NAME %in% d16_cities) |>
  st_union() |>
  st_make_valid()

tibble(
  district = c(15, 16),
  area_sqm = c(as.numeric(st_area(d15)), as.numeric(st_area(d16))),
  frac_of_downriver_cities = c(
    as.numeric(st_area(st_intersection(st_geometry(d15), target))) / as.numeric(st_area(target)),
    as.numeric(st_area(st_intersection(st_geometry(d16), target))) / as.numeric(st_area(target))
  )
)
```

95% of the Dearborn/Downriver cities that should be District 16 are instead
part of District 15 on file. This matches the issue exactly.

Checking the neighboring era files (`Michigan_093_to_097.geojson`,
`Michigan_098_to_102.geojson`, `Michigan_103_to_107.geojson`) shows the same
cities are correctly ~82-86% within District 16 there, confirming the bug
is isolated to `Michigan_089_to_092.geojson`.

## Fix

Move the Dearborn/Downriver cities (clipped to the existing District
15/16 combined area, so we don't take territory from neighboring districts
1 or 17) from District 15 to District 16, keeping the existing (correct)
Detroit panhandle piece as part of District 16 as well.

```{r}
region <- st_union(st_geometry(d15), st_geometry(d16)) |> st_make_valid()
target_clip <- st_intersection(target, region) |> st_make_valid()

new_d16_geom <- st_union(st_geometry(d16), target_clip) |> st_make_valid()
new_d15_geom <- st_difference(st_geometry(d15), target_clip) |> st_make_valid()

new_d15 <- st_set_geometry(d15, st_sfc(new_d15_geom, crs = st_crs(mi)))
new_d16 <- st_set_geometry(d16, st_sfc(new_d16_geom, crs = st_crs(mi)))
```

## Validation

Area is conserved and the two new shapes don't overlap:

```{r}
as.numeric(st_area(region))
as.numeric(st_area(new_d15)) + as.numeric(st_area(new_d16))
as.numeric(st_area(st_intersection(new_d15, new_d16)))
```

District 16 now covers essentially all of the Dearborn/Downriver cities
called for by the Census map, and District 15 essentially none of them:

```{r}
tibble(
  district = c(15, 16),
  frac_of_downriver_cities = c(
    as.numeric(st_area(st_intersection(st_geometry(new_d15), target))) / as.numeric(st_area(target)),
    as.numeric(st_area(st_intersection(st_geometry(new_d16), target))) / as.numeric(st_area(target))
  )
)
```

## Write out the corrected file

```{r}
out <- mi |>
  filter(!(district %in% c(15, 16))) |>
  bind_rows(new_d15, new_d16) |>
  arrange(district)

unlink(GEOJSON_FILE)
st_write(out, GEOJSON_FILE, append = FALSE)
```

## Visual validation against the Census map

As a quick sanity check, overlay the corrected District 15/16 shapes directly
on top of the Census Bureau's own scan of the 89th Congress Wayne County map
(the one attached to the issue). The scan has no embedded coordinates, so it
first has to be georeferenced: three small, distinctively-shaped
municipalities that are easy to pick out by eye in the scan -- Plymouth
city, Belleville city, and the Highland Park enclave -- and whose real-world
boundaries have not meaningfully changed since the 1960s, are used as
control points (their pixel bounding boxes in the scan, matched to their
`tigris` bounding boxes) to fit a simple pixel -> lon/lat transform.

```{r}
#| message: false
map_url <- "https://github.com/user-attachments/assets/a558df0c-39b7-4198-ab3b-647be0f32307"
map_file <- tempfile(fileext = ".png")
download.file(map_url, map_file, quiet = TRUE, mode = "wb")

img <- png::readPNG(map_file)
img_h <- dim(img)[1]
img_w <- dim(img)[2]

# Control points: pixel bounding-box corners (col, row) of Plymouth city,
# Belleville city, and the Highland Park enclave, read off the scan, matched
# to each city's real (NW, SE) bounding-box corner from `wayne_cousub`
# (confirmed via st_bbox() on each city, filtered to NAMELSAD == "... city"
# to avoid the same-named surrounding township).
gcp <- tibble(
  place = c("Plymouth NW", "Plymouth SE", "Belleville NW", "Belleville SE",
            "Highland Park NW", "Highland Park SE"),
  px    = c(208, 248, 190, 222, 657, 726),
  py    = c(191, 228, 467, 502, 140, 190),
  lon   = c(-83.48687, -83.44552, -83.49901, -83.47237, -83.12193, -83.07377),
  lat   = c(42.38445, 42.35850, 42.21386, 42.19006, 42.41816, 42.38808)
)

lon_fit <- lm(lon ~ px, data = gcp)
lat_fit <- lm(lat ~ py, data = gcp)

max(abs(residuals(lon_fit)))
max(abs(residuals(lat_fit)))

xmin <- predict(lon_fit, tibble(px = 0))
xmax <- predict(lon_fit, tibble(px = img_w))
ymax <- predict(lat_fit, tibble(py = 0))
ymin <- predict(lat_fit, tibble(py = img_h))
```

Residuals are well under 0.01 degrees (a few hundred meters) at county
scale, plenty precise for a visual check. Now plot the corrected shapes on
top of the georeferenced scan:

```{r}
#| fig-width: 9
#| fig-height: 9
ggplot() +
  annotation_raster(img, xmin, xmax, ymin, ymax) +
  geom_sf(data = new_d15, fill = NA, color = "red", linewidth = 1) +
  geom_sf(data = new_d16, fill = NA, color = "blue", linewidth = 1) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  labs(title = "Corrected District 15 (red) / District 16 (blue) vs. Census map") +
  theme_minimal()
```

The corrected District 16 (blue) now traces the Census map's compact
Dearborn/Downriver outline, and District 15 (red) wraps around it exactly
as shown on the map.
