---
title: "Fix Michigan CD 15/16 Boundary From Issue #24"
format: html
editor: source
---

## Overview

Benjamin Dasilva reported in [Issue #24](https://github.com/JeffreyBLewis/congressional-district-boundaries/issues/24)
that in `Michigan_089_to_092.geojson` (89th-92nd Congress, 1965-1972), the
suburban ("Downriver") portion of District 16 is incorrectly included in
District 15, and that the on-file District 16 shape only covers "the Detroit
portion of the district."

```{r}
#| message: false
library(tidyverse)
library(sf)
library(tigris)
options(tigris_use_cache = TRUE)
sf_use_s2(FALSE)
```

## Confirming the bug

```{r}
GEOJSON_FILE <- "../../GeoJson/Michigan_089_to_092.geojson"
mi <- st_read(GEOJSON_FILE, quiet = TRUE)
d15 <- mi |> filter(district == 15)
d16 <- mi |> filter(district == 16)
```

The Census Bureau's own map of Wayne County for the 89th Congress (attached
to the issue) shows District 16 as a compact area covering the city of
Dearborn and the inner Downriver suburbs -- Melvindale, Allen Park, Lincoln
Park, River Rouge, Ecorse, Southgate, and Wyandotte -- with District 15
wrapping around it to the west, north, and south (Canton, Nankin, Garden
City, Dearborn Heights, Inkster, Wayne, Romulus, Taylor, and the outer
Downriver communities further south like Trenton, Woodhaven, and Rockwood).

On file, `district == 16` is tiny (`{r} as.numeric(st_area(d16))` sq. m,
vs. `{r} as.numeric(st_area(d15))` for `district == 15`) and its shape does
not resemble Dearborn/Downriver at all. Comparing it to current Wayne
County, MI civil boundaries (which for these long-incorporated cities are
essentially unchanged since the 1960s) confirms what it actually is:

```{r}
wayne_cousub <- county_subdivisions(state = "MI", county = "Wayne", cb = TRUE, year = 2020) |>
  st_transform(st_crs(mi))

detroit <- wayne_cousub |> filter(NAME == "Detroit")
as.numeric(st_area(st_intersection(st_geometry(d16), st_geometry(detroit)))) /
  as.numeric(st_area(d16))
```

Essentially all (>99.9%) of the on-file District 16 shape sits inside the
City of Detroit -- specifically, its irregular southwestern ("Delray")
panhandle that pokes down between Melvindale, Dearborn, and River Rouge
along the Rouge River. That panhandle is geographically cut off from the
rest of Detroit by the surrounding suburbs, so it is plausible it really
was drawn into the surrounding 16th district. But the suburban cities that
should make up the bulk of District 16 -- Dearborn, Melvindale, Allen Park,
Lincoln Park, River Rouge, Ecorse, Southgate, and Wyandotte -- are missing
from it entirely:

```{r}
d16_cities <- c("Dearborn", "Melvindale", "Allen Park", "Lincoln Park",
                "River Rouge", "Ecorse", "Southgate", "Wyandotte")
target <- wayne_cousub |>
  filter(NAME %in% d16_cities) |>
  st_union() |>
  st_make_valid()

tibble(
  district = c(15, 16),
  area_sqm = c(as.numeric(st_area(d15)), as.numeric(st_area(d16))),
  frac_of_downriver_cities = c(
    as.numeric(st_area(st_intersection(st_geometry(d15), target))) / as.numeric(st_area(target)),
    as.numeric(st_area(st_intersection(st_geometry(d16), target))) / as.numeric(st_area(target))
  )
)
```

95% of the Dearborn/Downriver cities that should be District 16 are instead
part of District 15 on file. This matches the issue exactly.

Checking the neighboring era files (`Michigan_093_to_097.geojson`,
`Michigan_098_to_102.geojson`, `Michigan_103_to_107.geojson`) shows the same
cities are correctly ~82-86% within District 16 there, confirming the bug
is isolated to `Michigan_089_to_092.geojson`.

## Fix

Move the Dearborn/Downriver cities (clipped to the existing District
15/16 combined area, so we don't take territory from neighboring districts
1 or 17) from District 15 to District 16, keeping the existing (correct)
Detroit panhandle piece as part of District 16 as well.

```{r}
region <- st_union(st_geometry(d15), st_geometry(d16)) |> st_make_valid()
target_clip <- st_intersection(target, region) |> st_make_valid()

new_d16_geom <- st_union(st_geometry(d16), target_clip) |> st_make_valid()
new_d15_geom <- st_difference(st_geometry(d15), target_clip) |> st_make_valid()

new_d15 <- st_set_geometry(d15, st_sfc(new_d15_geom, crs = st_crs(mi)))
new_d16 <- st_set_geometry(d16, st_sfc(new_d16_geom, crs = st_crs(mi)))
```

## Validation

Area is conserved and the two new shapes don't overlap:

```{r}
as.numeric(st_area(region))
as.numeric(st_area(new_d15)) + as.numeric(st_area(new_d16))
as.numeric(st_area(st_intersection(new_d15, new_d16)))
```

District 16 now covers essentially all of the Dearborn/Downriver cities
called for by the Census map, and District 15 essentially none of them:

```{r}
tibble(
  district = c(15, 16),
  frac_of_downriver_cities = c(
    as.numeric(st_area(st_intersection(st_geometry(new_d15), target))) / as.numeric(st_area(target)),
    as.numeric(st_area(st_intersection(st_geometry(new_d16), target))) / as.numeric(st_area(target))
  )
)
```

## Write out the corrected file

```{r}
out <- mi |>
  filter(!(district %in% c(15, 16))) |>
  bind_rows(new_d15, new_d16) |>
  arrange(district)

unlink(GEOJSON_FILE)
st_write(out, GEOJSON_FILE, append = FALSE)
```
