---
title: "Fix Michigan CD 15/16 Boundary From Issue #24"
format: html
editor: source
---

## Overview

Benjamin Dasilva reported in [Issue #24](https://github.com/JeffreyBLewis/congressional-district-boundaries/issues/24)
that in `Michigan_089_to_092.geojson` (89th-92nd Congress, 1965-1972), the
suburban ("Downriver") portion of District 16 is incorrectly included in
District 15, and that the on-file District 16 shape only covers "the Detroit
portion of the district."

```{r}
#| message: false
library(tidyverse)
library(sf)
library(tigris)
library(readxl)
options(tigris_use_cache = TRUE)
sf_use_s2(FALSE)
```

## Confirming the bug

```{r}
GEOJSON_FILE <- "../../GeoJson/Michigan_089_to_092.geojson"
mi <- st_read(GEOJSON_FILE, quiet = TRUE)
d15 <- mi |> filter(district == 15)
d16 <- mi |> filter(district == 16)
```

On file, `district == 16` is tiny (`{r} as.numeric(st_area(d16))` sq. m,
vs. `{r} as.numeric(st_area(d15))` for `district == 15`). Comparing it to
current Wayne County, MI civil boundaries (which for these long-incorporated
cities are essentially unchanged since the 1960s) shows what it actually is:

```{r}
wayne_cousub <- county_subdivisions(state = "MI", county = "Wayne", cb = TRUE, year = 2020) |>
  st_transform(st_crs(mi))

detroit <- wayne_cousub |> filter(NAME == "Detroit")
as.numeric(st_area(st_intersection(st_geometry(d16), st_geometry(detroit)))) /
  as.numeric(st_area(d16))
```

Essentially all (>99.9%) of the on-file District 16 shape sits inside the
City of Detroit -- specifically, its irregular southwestern ("Delray")
panhandle that pokes down between Melvindale, Dearborn, and River Rouge
along the Rouge River. That panhandle is geographically cut off from the
rest of Detroit by the surrounding suburbs, so it plausibly really was
drawn into the surrounding 16th district. But essentially none of the
Downriver suburbs that should make up the bulk of District 16 are present.

## Determining the correct boundary

The Census Bureau's scanned map of Wayne County for the 89th Congress
(attached to the issue) shows District 16's approximate outline, but hand
tracing which side of a hand-drawn, 55-year-old scanned line a given city
falls on is error-prone -- an initial pass at this patch misread the line
around Melvindale/Allen Park/Lincoln Park from the scan alone and got it
wrong. The same UCLA project that produced this repository's source data
also published a per-state, per-era notes file documenting exactly which
counties/townships/cities were assigned to each district, citing the
actual 1964 Michigan apportionment law (Act 282) as its source. That's a
much more reliable, unambiguous reference than reading a scan, so it's
used here as ground truth:

```{r}
notes_url <- "https://cdmaps.polisci.ucla.edu/doc/89-92cc_Michigan_Notes.xlsx"
notes_file <- tempfile(fileext = ".xlsx")
download.file(notes_url, notes_file, quiet = TRUE, mode = "wb")

raw <- read_excel(notes_file, sheet = "Sheet1", col_names = FALSE)
notes <- raw |>
  filter(row_number() > which(raw[[1]] == "State")) |>
  set_names(c("state", "district", "county", "x4", "x5", "x6", "progress", "note")) |>
  filter(district %in% c("15", "16"), county == "Wayne") |>
  select(district, note)

notes |> pull(note) |> walk(\(x) {cat(strwrap(x, width = 90), sep = "\n"); cat("\n\n")})
```

District 15's Wayne County territory is: townships of Canton, Dearborn,
Huron, Nankin, Romulus, Sumpter, Taylor, Van Buren; cities of Allen Park,
Belleville, Dearborn Heights, Garden City, Lincoln Park, Melvindale, Wayne.

District 16's Wayne County territory is: townships of Brownstown and Grosse
Ile; cities of Dearborn, Ecorse, Gibraltar, River Rouge, Riverview,
Southgate, Trenton, Wyandotte; plus the part of the City of Detroit
described as the panhandle already correctly on file (confirmed above).

So Melvindale, Allen Park, and Lincoln Park -- despite sitting geographically
between Dearborn and the rest of Downriver -- are District 15, forming a
peninsula that pokes into what's otherwise District 16 territory. Woodhaven,
Flat Rock, and Rockwood aren't named explicitly (they incorporated out of
Brownstown Township in 1965-68, shortly after the 1964 law was written), but
nothing in the notes or the map shows any district line cutting them off
from the rest of the Downriver cluster, so they're treated as inheriting
Brownstown Township's District 16 assignment.

```{r}
d16_places <- c("Dearborn", "Ecorse", "River Rouge", "Southgate", "Wyandotte",
                "Riverview", "Trenton", "Gibraltar", "Woodhaven", "Flat Rock",
                "Rockwood", "Brownstown", "Grosse Ile")
target <- wayne_cousub |>
  filter(NAME %in% d16_places) |>
  st_union() |>
  st_make_valid()

tibble(
  district = c(15, 16),
  area_sqm = c(as.numeric(st_area(d15)), as.numeric(st_area(d16))),
  frac_of_target_cities = c(
    as.numeric(st_area(st_intersection(st_geometry(d15), target))) / as.numeric(st_area(target)),
    as.numeric(st_area(st_intersection(st_geometry(d16), target))) / as.numeric(st_area(target))
  )
)
```

96% of the cities/townships that should be District 16 are instead part of
District 15 on file, confirming the bug.

Checking the neighboring era files (`Michigan_093_to_097.geojson`,
`Michigan_098_to_102.geojson`, `Michigan_103_to_107.geojson`) shows their
`district == 15`/`16` boundaries were redrawn at each subsequent
redistricting (Michigan's district lines here were not static), so they
aren't used to infer the 89-92 boundary -- the notes file is the only
source used for that.

## Fix

Move the correct District 16 cities/townships (clipped to the existing
District 15/16 combined area, so we don't take territory from neighboring
districts 1 or 17) from District 15 to District 16, keeping the existing
(correct) Detroit panhandle piece as part of District 16 as well.

```{r}
region <- st_union(st_geometry(d15), st_geometry(d16)) |> st_make_valid()
target_clip <- st_intersection(target, region) |> st_make_valid()

new_d16_geom <- st_union(st_geometry(d16), target_clip) |> st_make_valid()
new_d15_geom <- st_difference(st_geometry(d15), target_clip) |> st_make_valid()

new_d15 <- st_set_geometry(d15, st_sfc(new_d15_geom, crs = st_crs(mi)))
new_d16 <- st_set_geometry(d16, st_sfc(new_d16_geom, crs = st_crs(mi)))
```

## Validation

Area is conserved and the two new shapes don't overlap:

```{r}
as.numeric(st_area(region))
as.numeric(st_area(new_d15)) + as.numeric(st_area(new_d16))
as.numeric(st_area(st_intersection(new_d15, new_d16)))
```

District 16 now covers essentially all of its target cities/townships, and
District 15 essentially none of them:

```{r}
tibble(
  district = c(15, 16),
  frac_of_target_cities = c(
    as.numeric(st_area(st_intersection(st_geometry(new_d15), target))) / as.numeric(st_area(target)),
    as.numeric(st_area(st_intersection(st_geometry(new_d16), target))) / as.numeric(st_area(target))
  )
)
```

And each individual city/township lands in the district the notes call for
(cross-checking Melvindale, Allen Park, and Lincoln Park stayed in District
15 rather than following their Downriver neighbors into District 16):

```{r}
check_places <- c("Melvindale", "Allen Park", "Lincoln Park", "Dearborn",
                   "Ecorse", "Wyandotte", "Grosse Ile", "Rockwood")
map_dfr(check_places, function(nm) {
  cg <- wayne_cousub |> filter(NAME == nm) |> st_geometry() |> st_union()
  tibble(
    place = nm,
    frac_in_d15 = as.numeric(st_area(st_intersection(cg, st_geometry(new_d15)))) / as.numeric(st_area(cg)),
    frac_in_d16 = as.numeric(st_area(st_intersection(cg, st_geometry(new_d16)))) / as.numeric(st_area(cg))
  )
})
```

## Write out the corrected file

```{r}
out <- mi |>
  filter(!(district %in% c(15, 16))) |>
  bind_rows(new_d15, new_d16) |>
  arrange(district)

unlink(GEOJSON_FILE)
st_write(out, GEOJSON_FILE, append = FALSE)
```

## Visual validation against the Census map

As a quick sanity check, overlay the corrected District 15/16 shapes
directly on top of the Census Bureau's own scan of the 89th Congress Wayne
County map (the one attached to the issue). The scan has no embedded
coordinates, so it first has to be georeferenced: three small,
distinctively-shaped municipalities that are easy to pick out by eye in the
scan -- Plymouth city, Belleville city, and the Highland Park enclave -- and
whose real-world boundaries have not meaningfully changed since the 1960s,
are used as control points (their pixel bounding boxes in the scan, matched
to their `tigris` bounding boxes) to fit a simple pixel -> lon/lat
transform.

```{r}
#| message: false
map_url <- "https://github.com/user-attachments/assets/a558df0c-39b7-4198-ab3b-647be0f32307"
map_file <- tempfile(fileext = ".png")
download.file(map_url, map_file, quiet = TRUE, mode = "wb")

img <- png::readPNG(map_file)
img_h <- dim(img)[1]
img_w <- dim(img)[2]

# Control points: pixel bounding-box corners (col, row) of Plymouth city,
# Belleville city, and the Highland Park enclave, read off the scan, matched
# to each city's real (NW, SE) bounding-box corner from `wayne_cousub`
# (confirmed via st_bbox() on each city, filtered to NAMELSAD == "... city"
# to avoid the same-named surrounding township).
gcp <- tibble(
  place = c("Plymouth NW", "Plymouth SE", "Belleville NW", "Belleville SE",
            "Highland Park NW", "Highland Park SE"),
  px    = c(208, 248, 190, 222, 657, 726),
  py    = c(191, 228, 467, 502, 140, 190),
  lon   = c(-83.48687, -83.44552, -83.49901, -83.47237, -83.12193, -83.07377),
  lat   = c(42.38445, 42.35850, 42.21386, 42.19006, 42.41816, 42.38808)
)

lon_fit <- lm(lon ~ px, data = gcp)
lat_fit <- lm(lat ~ py, data = gcp)

max(abs(residuals(lon_fit)))
max(abs(residuals(lat_fit)))

xmin <- predict(lon_fit, tibble(px = 0))
xmax <- predict(lon_fit, tibble(px = img_w))
ymax <- predict(lat_fit, tibble(py = 0))
ymin <- predict(lat_fit, tibble(py = img_h))
```

Residuals are well under 0.01 degrees (a few hundred meters) at county
scale, plenty precise for a visual check. Now plot the corrected shapes on
top of the georeferenced scan:

```{r}
#| fig-width: 9
#| fig-height: 9
ggplot() +
  annotation_raster(img, xmin, xmax, ymin, ymax) +
  geom_sf(data = new_d15, fill = NA, color = "red", linewidth = 1) +
  geom_sf(data = new_d16, fill = NA, color = "blue", linewidth = 1) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  labs(title = "Corrected District 15 (red) / District 16 (blue) vs. Census map") +
  theme_minimal()
```

And zoomed in on Allen Park/Melvindale/Lincoln Park, to check that peninsula
specifically against the map:

```{r}
#| fig-width: 8
#| fig-height: 8
ggplot() +
  annotation_raster(img, xmin, xmax, ymin, ymax) +
  geom_sf(data = new_d15, fill = NA, color = "red", linewidth = 1.2) +
  geom_sf(data = new_d16, fill = NA, color = "blue", linewidth = 1.2) +
  coord_sf(xlim = c(-83.22, -83.12), ylim = c(42.25, 42.35), expand = FALSE) +
  labs(title = "Zoom: Allen Park / Melvindale / Lincoln Park / Ecorse") +
  theme_minimal()
```

The corrected District 16 (blue) traces the Census map's Downriver outline,
including the full extent south to Rockwood/Flat Rock, while District 15
(red) keeps its Allen Park/Melvindale/Lincoln Park peninsula intact.
