---
title: "Revert an Unsupported Rhode Island Congress Split From Issue #19"
format: html
editor: source
---

## Overview

[Issue #19](https://github.com/JeffreyBLewis/congressional-district-boundaries/issues/19)
reported that Rhode Island's 114th- and 115th-Congress shapefiles disagreed
noticeably north of Providence, even though Census's own redistricting
page shows Rhode Island was *not* one of the states redistricted between
those two Congresses. Follow-up investigation on the issue concluded this
was a data artifact in this repository's original (2016-vintage) pull for
the 113th-114th Congress range, not a real boundary change -- current
Census data matches `Rhode Island_113_to_114.geojson` closely for the rest
of the period.

That conclusion was implemented in commit `b89f2bd` (PR #21, "Fix ri",
merged 2026-06-19), which added two new files,
`Rhode Island_113_to_117.geojson` and `Rhode Island_118_to_119.geojson`,
apparently intending to also split off a second, smaller change between
the 117th and 118th Congress. But it never removed the two files it was
meant to replace (`Rhode Island_113_to_114.geojson` and
`Rhode Island_115_to_119.geojson`), and the new files were written with
raw Census/tigris column names (`STATEFP`, `CD119FP`, `STARTCONG`
uppercase, etc.) instead of this repository's canonical schema
(`statename`, `district`, `startcong`, `endcong`, `id`, ...). The result:
Rhode Island has had four overlapping, duplicate-coverage files sitting in
`GeoJson/` since June, two of which are missing the columns the rest of
the pipeline depends on -- which is what surfaced this while testing an
unrelated shapefile-export workflow.

```{r}
#| message: false
library(tidyverse)
library(sf)
library(tigris)
options(tigris_use_cache = TRUE)
source("../../R/dedup.R")
```

## Confirming the schema break

```{r}
bad1 <- st_read("../../GeoJson/Rhode Island_113_to_117.geojson", quiet = TRUE)
bad2 <- st_read("../../GeoJson/Rhode Island_118_to_119.geojson", quiet = TRUE)
names(bad1)
names(bad2)
```

Neither has a `statename`, `startcong`, or `endcong` column in the
lowercase form every other file in this collection uses -- `bad1` has no
Congress-range columns at all. Loading the full collection this way is
exactly what makes `1:max(congressional_district_sf$endcong)` in
`_targets.R` fail with `NA/NaN argument`: these four rows contribute `NA`
to `endcong`.

## Checking whether the 117th-118th split is real

`R/dedup.R`'s `dedup_districts()` is this project's own, already-validated
rule for exactly this question: it merges consecutive Congresses for a
district into one shape whenever consecutive-vintage geometries overlap
90%+ -- more precisely, at or above a 0.99 intersection-over-union
threshold -- and keeps a break only when overlap drops below that. It's
the same rule that built the rest of the collection. Reapplying it here
with fresh data settles the question directly instead of relying on
memory of what the original PR compared.

```{r}
#| message: false
#| results: hide
baseline <- st_read("../../GeoJson/Rhode Island_113_to_114.geojson", quiet = TRUE) |>
  st_transform(4269) |>
  transmute(STATENAME = statename, DISTRICT = as.character(district),
            STARTCONG = 113, ENDCONG = 114, NOTE = note, geometry = geometry)

pull_year <- function(year, cong) {
  d <- congressional_districts(year = year, cb = TRUE, resolution = "500k", state = "44")
  cn <- grep("^CD[0-9]+FP$", names(d), value = TRUE)[1]
  d |> st_transform(4269) |>
    transmute(STATENAME = "Rhode Island", DISTRICT = .data[[cn]],
              STARTCONG = cong, ENDCONG = cong,
              NOTE = sprintf("From `tigris` using year = %d.", year),
              geometry = geometry)
}

all_ri <- bind_rows(
  baseline,
  pull_year(2017, 115),
  pull_year(2019, 116),
  pull_year(2019, 117),
  pull_year(2022, 118),
  pull_year(2024, 119)
) |>
  mutate(DISTRICT = as.numeric(DISTRICT))
```

```{r}
detail <- all_ri |>
  arrange(STATENAME, DISTRICT, STARTCONG) |>
  mutate(pchng = map2_dbl(geometry, lag(geometry),
                           \(g1, g2) if (is.null(g2)) NA else
                             as.numeric(st_area(st_intersection(g1, g2))) /
                             as.numeric(st_area(st_union(g1, g2)))),
         .by = c(STATENAME, DISTRICT))
st_drop_geometry(detail)[, c("DISTRICT", "STARTCONG", "ENDCONG", "pchng")]
```

The only transition below the 0.99 threshold is 113/114 &rarr; 115 (0.968
for District 1) -- the artifact issue #19 already identified. Every other
transition, including 117 &rarr; 118 (0.994 and 0.999), stays above it.
Running the collection's own merge rule on this confirms it directly:

```{r}
result <- dedup_districts(all_ri, .thresh = 0.99)
st_drop_geometry(result)[, c("STATENAME", "DISTRICT", "STARTCONG", "ENDCONG")]
```

Two ranges fall out: 113-114 and 115-119 -- exactly what
`Rhode Island_113_to_114.geojson` and `Rhode Island_115_to_119.geojson`
already encode. There is no support for a break at 117/118. (As a further
check, the geometry actually written into `Rhode Island_113_to_117.geojson`
and `Rhode Island_118_to_119.geojson` turns out to be identical between
the two files to begin with -- consistent with there being no real change
between them, but then there was no reason to split them into two files
at all.)

## Fix

The two files being replaced by `b89f2bd` were never removed, and the two
that replaced them don't match this repository's schema or its own
change-detection rule. The fix is to remove the two later files and keep
the originals, which is what the data already supports.

```{r}
file.remove("../../GeoJson/Rhode Island_113_to_117.geojson")
file.remove("../../GeoJson/Rhode Island_118_to_119.geojson")
```

## Validate

```{r}
files <- list.files("../../GeoJson", pattern = "Rhode Island", full.names = TRUE)
basename(files)

ri_check <- st_read("../../GeoJson/Rhode Island_113_to_114.geojson", quiet = TRUE) |>
  bind_rows(st_read("../../GeoJson/Rhode Island_115_to_119.geojson", quiet = TRUE))
st_drop_geometry(ri_check)[, c("statename", "district", "startcong", "endcong")]
all(st_is_valid(ri_check))
```

Rhode Island is back to two files, both with the canonical schema, and
together they cover the 113th through 119th Congress with no gaps or
overlaps. The pipeline's `congresses <- 1:max(congressional_district_sf$endcong)`
step, which failed on the stray `NA`s from the malformed files, now has
none to trip over.
