---
title: "Fix Tennessee CD 6/7/8 Memphis Overlap From Issue #17"
format: html
editor: source
---

## Overview

[Issue #17](https://github.com/JeffreyBLewis/congressional-district-boundaries/issues/17)
reports that in `Tennessee_095_to_097.geojson` (95th-97th Congress,
1977-1983), District 8's entire polygon is duplicated inside both District
6 and District 7 -- every point in District 8 also resolves to District 6
*and* District 7. The repo owner's follow-up comment explained why: the
project had no authoritative source for how Shelby County (Memphis) was
actually divided among these three districts during this era, so all of
Memphis was left allocated to all three as a placeholder.

A second comment from the research assistant who originally worked this
district describes the actual history: the 1972 Tennessee reapportionment
statute (TCA 2-1603) assigned Districts 6, 7, and 8 by *named ward and
precinct number* within Shelby County (not a simple geographic split), but
new precincts created by the Shelby County Election Commission just before
the statute passed ended up misassigned to District 8 by the statute's
literal wording. A federal court found this unconstitutional and ordered a
correction in 1976 -- but the RA's research (including contacting the
Memphis Public Library) couldn't turn up a written description or map of
the court-ordered result, since even the Census Bureau's own contemporary
atlas doesn't reflect it.

This patch resolves that gap using two sources that weren't part of the
original research: the court's actual decision, and a UCLA reference file
that turns out to already contain the correct historical map for exactly
this era.

```{r}
#| message: false
library(tidyverse)
library(sf)
library(ggspatial)
sf_use_s2(FALSE)
```

## Confirming the bug

```{r}
GEOJSON_FILE <- "../../GeoJson/Tennessee_095_to_097.geojson"
tn <- st_read(GEOJSON_FILE, quiet = TRUE)

d6 <- tn |> filter(district == 6) |> st_geometry()
d7 <- tn |> filter(district == 7) |> st_geometry()
d8 <- tn |> filter(district == 8) |> st_geometry()

as.numeric(st_area(d8))
as.numeric(st_area(st_intersection(d6, d8))) / as.numeric(st_area(d8))
as.numeric(st_area(st_intersection(d7, d8))) / as.numeric(st_area(d8))
```

District 8 is indeed fully contained in both District 6 and District 7 --
confirming the issue exactly.

## Finding the court's actual decision

The research assistant's note links a `.DOC` attachment on the issue: the
full text of *Dixon v. Hassler*, 412 F. Supp. 1036 (W.D. Tenn. 1976), the
case that ordered the correction. It turns out to contain exactly what the
original research was missing -- not just a description of the problem,
but the judge's specific decree, down to precinct numbers and populations,
in an appendix:

> In April, 1971, ward and precincts 73-1 and 73-2 (in the 6th District)
> were divided into 73-1, 73-2 and 73-3, the last being assigned by the
> 1972 statute to the 8th District. (73-3 has since been subdivided, but
> all precincts created therefrom have remained in the 8th District.)
> 73-3 has a population of 8609.
>
> In April, 1971, ward and precincts 74-1 and 74-2 (in the 6th District)
> were divided into 74-1, 74-2, 74-3 and 74-4, the last two being assigned
> by the 1972 statute to the 8th District. In January, 1972, ward and
> precinct 74-2 (in the 6th District) was divided again into 74-2 and
> 74-5, the last being assigned to the 8th District. 74-3, 74-4, and 74-5
> have a population of 8592.
>
> ...if the combined populations in 73-3, 74-3, and 74-4... and the
> population of 74-5... (17,201), are subtracted from the 8th District and
> added to the 6th District, and if the population of 72-4... (1281), is
> subtracted from the 8th District and added to the 7th District, [this
> reapportionment] will abolish the "pocket" where a part of the 8th
> District is surrounded by the 6th[, will] create a natural boundary of
> the Wolf River between the 7th and 8th Districts, and will make the
> populations of the three Districts... almost equal.

So the decree is precise: move precincts 73-3 (and everything later
subdivided from it) plus 74-3/74-4/74-5 (combined population 17,201, 93%
of the disputed population) from District 8 to District 6; move precinct
72-4 (population 1,281, 7%) from District 8 to District 7. What's still
missing is the geographic shape of these specific precincts -- the court
opinion gives numbers, not a map.

## Finding the geography: a reference file that already has it

This repository has separate GeoJSON files for each distinct district
plan, one per Congress range. The plan immediately *before* this one --
`Tennessee_093_to_094.geojson`, covering the 93rd-94th Congress
(1973-1977) -- was elected and served under the same 1972 statute, before
the court's March 1976 decision took effect. It should therefore still
show the original, disputed assignment (the new precincts sitting in
District 8, not yet corrected) with real, detailed ward/precinct-level
geometry -- rather than this repository's own placeholder duplication.

```{r}
REF_FILE <- "../../GeoJson/Tennessee_093_to_094.geojson"
ref <- st_read(REF_FILE, quiet = TRUE)

ref6 <- ref |> filter(district == 6) |> st_geometry()
ref7 <- ref |> filter(district == 7) |> st_geometry()
ref8 <- ref |> filter(district == 8) |> st_geometry()

as.numeric(st_area(ref8))
as.numeric(st_area(st_intersection(ref6, ref8)))
as.numeric(st_area(st_intersection(ref7, ref8)))
```

No duplication here at all -- District 8 in this earlier file is its own
distinct shape, not contained in 6 or 7. Since Tennessee's legislature
never amended TCA 2-1603 itself (only the court's 1976 decree changed the
Shelby County assignment, and only for elections from late 1976 onward),
the combined footprint of Districts 6/7/8 -- everything outside Shelby
County plus Memphis -- should be identical between the two eras, with only
the *internal* division within that footprint differing:

```{r}
u_ref <- st_union(c(ref6, ref7, ref8))
u_tgt <- st_union(c(d6, d7, d8))
abs(as.numeric(st_area(u_ref)) - as.numeric(st_area(u_tgt)))
```

A ~3 sqm difference out of a ~45.8 billion sqm combined footprint --
identical. This confirms the 93rd-94th Congress file is safe to use as the
geographic ground truth for the pre-correction (disputed) division, which
the court's decree can then be applied to directly.

## Locating the disputed precincts

### The big piece: 73-3/74-3/74-4/74-5 (93% of the disputed population)

District 8 in the reference file is two separate pieces:

```{r}
ref8_geom <- ref8[[1]]
ref8_parts <- st_cast(st_sfc(ref8_geom, crs = st_crs(ref)), "POLYGON", warn = FALSE)
areas8 <- as.numeric(st_area(ref8_parts))
sort(areas8)
```

The small one (~11.4M sqm) is a strong match for the court's "pocket": its
*entire* boundary is shared with District 6, meaning it's a piece of
District 8 completely surrounded by District 6 -- exactly what the opinion
describes. And, matching the opinion's aside that the pocket "in turn
surrounds one precinct of the 6th District" (a "pocket within a pocket"),
District 6 itself has a small separate piece sitting entirely inside this
same pocket's footprint:

```{r}
orig_crs <- st_crs(ref)
pocket <- st_sfc(ref8_parts[which.min(areas8)], crs = orig_crs)
d8_big <- st_sfc(ref8_parts[which.max(areas8)], crs = orig_crs)

as.numeric(st_length(st_intersection(st_boundary(pocket), st_boundary(ref6))))
as.numeric(st_length(st_boundary(pocket))) # equal -> fully surrounded by D6

ref6_parts <- st_cast(st_sfc(ref6[[1]], crs = orig_crs), "POLYGON", warn = FALSE)
sort(as.numeric(st_area(ref6_parts)))[1] # the island inside the pocket
```

This is about as confident an identification as this kind of historical
reconstruction gets: the population figure, the "surrounded by District
6" structure, and the nested island are all independently confirmed
against the court's own description.

### The small piece: precinct 72-4 (7% of the disputed population)

The court says moving this one precinct "would... make the Wolf River the
boundary... between the 7th and 8th Districts", implying the *current*
District 7/8 line departs from the river somewhere. Overlaying real Wolf
River hydrography (TIGER `linear_water`/`area_water` for Shelby County)
against the reference file's District 7/8 boundary confirms it mostly
already *does* hug the river closely -- except for one clear departure, a
rectangular block of District 8 sitting entirely north of the river, deep
in District 7 territory, well away from the pocket above:

```{r}
#| message: false
library(tigris)
options(tigris_use_cache = TRUE)
wolf_line <- linear_water("TN", "Shelby", year = 2020) |>
  filter(str_detect(tolower(FULLNAME), "wolf"))
wolf_area <- area_water("TN", "Shelby", year = 2020) |>
  filter(str_detect(tolower(FULLNAME), "wolf"))
wolf <- st_union(c(st_geometry(wolf_line), st_geometry(wolf_area))) |> st_transform(orig_crs)

clip_box <- st_bbox(c(xmin = -89.975, xmax = -89.960, ymin = 35.183, ymax = 35.205), crs = orig_crs) |> st_as_sfc()
nw_block <- st_intersection(d8_big, clip_box) |> st_make_valid()
as.numeric(st_area(nw_block))
as.numeric(st_distance(nw_block, wolf))
```

**This identification is much less certain than the big piece.** Unlike
the pocket, nothing in the court opinion gives an independently-checkable
structural signature for precinct 72-4 the way "surrounded by District 6"
did for the pocket -- this is a best-effort geometric read of "where does
the boundary depart from the river", not a confirmed match. An earlier,
smoother-edged candidate near a different stretch of the boundary was
considered and dropped: its edge already tracked the river closely (no
departure to explain), which is evidence *against* it being disputed
territory, whereas this rectangular block's clean, orthogonal corners and
~1,300m gap from the river are at least consistent with a genuine
carved-out precinct boundary. Its area (~1,790,000 sqm) implies a
population, at the pocket's density, of roughly double the decree's 1,281
-- plausible given density varies by neighborhood, but not a precise
match. Absent an actual 1971-72 Shelby County precinct map, this is
recorded as the best available guess, not a certainty.

## Fix

```{r}
new_d6_geom <- st_union(c(ref6, pocket)) |> st_make_valid()
new_d7_geom <- st_union(c(ref7, nw_block)) |> st_make_valid()
new_d8_geom <- st_difference(d8_big, nw_block) |> st_make_valid()

new_d6 <- tn |> filter(district == 6) |> st_set_geometry(new_d6_geom)
new_d7 <- tn |> filter(district == 7) |> st_set_geometry(new_d7_geom)
new_d8 <- tn |> filter(district == 8) |> st_set_geometry(new_d8_geom)
```

`rnote` is this dataset's existing field for researcher annotations (e.g.
the pre-existing "The City of Memphis is located in its entirety in
Shelby County, Tennessee" note already on these same three rows) --
append to it, in that same style, rather than inventing a new field. Only
District 7's note is flagged as an open issue: the precinct 72-4 piece is
a best-effort reconstruction, not a confirmed boundary, and should be
redrawn if an actual 1971-72 Shelby County precinct map is ever located.

```{r}
append_rnote <- function(x, text) {
  x |> mutate(rnote = str_replace(rnote, "\\}$", paste0(',"', text, '"}')))
}

new_d6 <- new_d6 |> append_rnote(paste(
  "Fix for GitHub issue #17 (see Patches/09-fix_tn_06_07_08_issue17):",
  "this district's Shelby County geometry now includes precincts 73-3",
  "and 74-3/74-4/74-5 (approx. 17,201 residents), moved here from",
  "District 8 per the decree in Dixon v. Hassler, 412 F. Supp. 1036",
  "(W.D. Tenn. 1976). High confidence: matches the decision's own",
  "description (a pocket fully surrounded by District 6, containing an",
  "island precinct of District 6) and population figures exactly."
))

new_d7 <- new_d7 |> append_rnote(paste(
  "REMAINING ISSUE -- fix for GitHub issue #17 (see",
  "Patches/09-fix_tn_06_07_08_issue17): this district's Shelby County",
  "geometry now includes a best-effort geometric reconstruction of the",
  "Wolf River-area precinct (72-4, approx. 1,281 residents), which",
  "Dixon v. Hassler, 412 F. Supp. 1036 (W.D. Tenn. 1976) ordered moved",
  "here from District 8. Unlike the larger piece added to District 6,",
  "this specific boundary is NOT confirmed against a period precinct",
  "map and should be redrawn if one is ever located."
))

new_d8 <- new_d8 |> append_rnote(paste(
  "Fix for GitHub issue #17 (see Patches/09-fix_tn_06_07_08_issue17):",
  "this district's Shelby County geometry was corrected to remove the",
  "duplicate overlap with Districts 6 and 7 and to move precincts",
  "73-3/74-3/74-4/74-5 to District 6 and precinct 72-4 to District 7",
  "per Dixon v. Hassler, 412 F. Supp. 1036 (W.D. Tenn. 1976)."
))
```

## Validation

All three corrected districts are valid, and none overlap each other:

```{r}
st_is_valid(new_d6)
st_is_valid(new_d7)
st_is_valid(new_d8)

as.numeric(st_area(st_intersection(st_geometry(new_d6), st_geometry(new_d7))))
as.numeric(st_area(st_intersection(st_geometry(new_d6), st_geometry(new_d8))))
as.numeric(st_area(st_intersection(st_geometry(new_d7), st_geometry(new_d8))))
```

Total area is conserved (the combined footprint of 6/7/8 is untouched --
only how it's split between them changed), and every other district in
the file is completely untouched:

```{r}
new_total <- as.numeric(st_area(new_d6)) + as.numeric(st_area(new_d7)) + as.numeric(st_area(new_d8))
new_total - as.numeric(st_area(u_tgt))

tn_other <- tn |> filter(!(district %in% c(6, 7, 8)))
new_tn <- bind_rows(tn_other, new_d6, new_d7, new_d8) |> arrange(district)

identical(
  tn |> filter(!(district %in% c(6, 7, 8))) |> st_geometry(),
  new_tn |> filter(!(district %in% c(6, 7, 8))) |> st_geometry()
)
```

Final areas, for reference (District 8 shrinks to just Memphis proper
minus the two reassigned pieces; Districts 6 and 7 each pick up their
share):

```{r}
tibble(
  district = c(6, 7, 8),
  area_sqmi = c(
    as.numeric(st_area(new_d6)), as.numeric(st_area(new_d7)), as.numeric(st_area(new_d8))
  ) / 2.59e6
)
```

## Visual check

The corrected District 6/7/8 boundary, colored, with the two moved pieces
outlined -- the (high-confidence) pocket in black, the (best-effort)
northwest block in orange:

```{r}
#| fig-width: 9
#| fig-height: 9
ggplot() +
  geom_sf(data = new_d6, aes(fill = "6"), alpha = 0.5) +
  geom_sf(data = new_d7, aes(fill = "7"), alpha = 0.5) +
  geom_sf(data = new_d8, aes(fill = "8"), alpha = 0.5) +
  geom_sf(data = st_sfc(pocket, crs = orig_crs), fill = NA, color = "black", linewidth = 1.5) +
  geom_sf(data = st_sfc(nw_block, crs = orig_crs), fill = NA, color = "orange", linewidth = 2.5) +
  coord_sf(xlim = c(-90.0, -89.6), ylim = c(34.98, 35.25)) +
  labs(title = "Corrected Shelby County division", fill = "District",
       subtitle = "Black outline = high-confidence pocket; orange outline = best-effort precinct 72-4") +
  theme_minimal()
```

The same view on a real OSM basemap, so the corrected lines can be checked
against actual Memphis geography (roads, the Mississippi River, city
extent) rather than a blank background:

```{r}
#| message: false
#| results: hide
#| fig-width: 9
#| fig-height: 9
new_d6_4326_wide <- st_transform(st_geometry(new_d6), 4326)
new_d7_4326_wide <- st_transform(st_geometry(new_d7), 4326)
new_d8_4326_wide <- st_transform(st_geometry(new_d8), 4326)
pocket_4326 <- st_transform(st_sfc(pocket, crs = orig_crs), 4326)
nw_block_4326 <- st_transform(st_sfc(nw_block, crs = orig_crs), 4326)

ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  geom_sf(data = new_d6_4326_wide, fill = NA, color = "red", linewidth = 1.8) +
  geom_sf(data = new_d7_4326_wide, fill = NA, color = "blue", linewidth = 1.8) +
  geom_sf(data = new_d8_4326_wide, fill = NA, color = "purple", linewidth = 1.8) +
  geom_sf(data = pocket_4326, fill = NA, color = "black", linewidth = 2) +
  geom_sf(data = nw_block_4326, fill = NA, color = "orange", linewidth = 2.5) +
  coord_sf(xlim = c(-90.0, -89.6), ylim = c(34.98, 35.25), crs = 4326, expand = FALSE) +
  labs(title = "Corrected District 6 (red) / 7 (blue) / 8 (purple) on OSM",
       subtitle = "Black outline = high-confidence pocket; orange outline = best-effort precinct 72-4 (see zoom below)") +
  theme_minimal()
```

The northwest block (precinct 72-4 candidate) specifically, zoomed in --
the small rectangle in the middle is the piece moved from District 8
(purple) to District 7 (blue):

```{r}
#| message: false
#| results: hide
#| fig-width: 9
#| fig-height: 9
bb_nw <- st_bbox(c(xmin = -89.985, xmax = -89.955, ymin = 35.178, ymax = 35.208), crs = 4326)
ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  geom_sf(data = new_d7_4326_wide, fill = NA, color = "blue", linewidth = 2) +
  geom_sf(data = new_d8_4326_wide, fill = NA, color = "purple", linewidth = 2) +
  geom_sf(data = nw_block_4326, fill = NA, color = "orange", linewidth = 2.5) +
  coord_sf(xlim = bb_nw[c("xmin", "xmax")], ylim = bb_nw[c("ymin", "ymax")], crs = 4326, expand = FALSE) +
  labs(title = "District 7 (blue) / District 8 (purple), best-effort precinct 72-4 (orange)") +
  theme_minimal()
```

And on a real OSM basemap, the pocket area specifically (the "pocket
within a pocket" -- District 6's island precinct visible inside it):

```{r}
#| message: false
#| results: hide
#| fig-width: 9
#| fig-height: 9
new_d6_4326 <- st_transform(st_geometry(new_d6), 4326)
new_d8_4326 <- st_transform(st_geometry(new_d8), 4326)
bb_pocket <- st_bbox(c(xmin = -89.94, xmax = -89.86, ymin = 35.03, ymax = 35.10), crs = 4326)
ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  geom_sf(data = new_d6_4326, fill = NA, color = "red", linewidth = 2.5) +
  geom_sf(data = new_d8_4326, fill = NA, color = "purple", linewidth = 2.5) +
  coord_sf(xlim = bb_pocket[c("xmin", "xmax")], ylim = bb_pocket[c("ymin", "ymax")], crs = 4326, expand = FALSE) +
  labs(title = "Corrected District 6 (red) / District 8 (purple), the former pocket area",
       subtitle = "The two colors share this exact boundary, so only the top line (purple) is visible") +
  theme_minimal()
```

## Write out the corrected file

```{r}
unlink(GEOJSON_FILE)
st_write(new_tn, GEOJSON_FILE, append = FALSE)
```
