---
title: "Fix At-Large District Overlap From Issue #22"
format: html
editor: source
---

## Overview

@cgrabois reported in [Issue #22](https://github.com/JeffreyBLewis/congressional-district-boundaries/issues/22)
that in several shapefiles, "District 1" appears to equal the full state
boundary, almost entirely overlapping "District 2", for South Dakota
(93rd-97th Congress), Montana (93rd-98th and 99th-102nd Congress), and (per a
follow-up comment) Montana, North Dakota, and South Dakota in the 92nd
Congress.

Investigating, the individual `district == 1` and `district == 2` shapes in
the relevant per-era GeoJSON files (e.g. `South Dakota_093_to_097.geojson`)
are already correct and non-overlapping: their areas sum to the full state
area with essentially zero intersection. The actual bug is that each of these
three states also has a `district == 0` ("at-large") shape -- a single
polygon equal to the entire state, used for congresses when the state had
only one, statewide district:

  - `South Dakota_051_to_119.geojson`
  - `Montana_051_to_117.geojson`
  - `North Dakota_051_to_119.geojson`

Each of these files has `startcong`/`endcong` values that span the state's
*entire* congressional history, including congresses when the state actually
had two (or three) separate districts. So for, say, the 93rd Congress, South
Dakota has three shapes on file: the correct District 1, the correct District
2, *and* a spurious full-state "district 0" shape that duplicates/overlaps
both of them. Downstream code that combines these files by congress (like
this repo's own `_targets.R` pipeline, or the shapefiles the reporter built)
ends up with a full-state polygon sitting on top of the real districts --
which is what produces the "District 1 = whole state" appearance.

This is a metadata bug, not a geometry bug: the at-large polygon itself is
correct (it really is the full state, for the congresses when the state
really was at-large). The fix is to split each `district == 0` row's
congress range so it no longer overlaps the congresses covered by the
state's other (multi-district) files, re-using the same at-large geometry
for each split range.

```{r}
#| message: false
library(tidyverse)
library(sf)
sf_use_s2(FALSE)
```

## Determine the correct at-large ranges

For each state, the multi-district files already on file tell us exactly
which congresses were *not* at-large. The at-large row should cover
everything else within its original range.

```{r}
GEOJSON_DIR <- "../../GeoJson"

district_files <- function(statename) {
  list.files(GEOJSON_DIR,
             pattern = paste0("^", statename, "_[0-9]+_to_[0-9]+\\.geojson$"),
             full.names = TRUE)
}

congress_coverage <- function(statename) {
  map_dfr(district_files(statename), function(f) {
    st_read(f, quiet = TRUE) |>
      st_drop_geometry() |>
      distinct(district, startcong, endcong) |>
      mutate(file = basename(f))
  }) |>
    arrange(startcong)
}

congress_coverage("South Dakota")
congress_coverage("Montana")
congress_coverage("North Dakota")
```

The `district == 0` (at-large) row's original range, and the non-at-large
ranges it needs to avoid, for each state:

  - **South Dakota**: at-large file covers 51-119; districted 63-97 (no gaps).
    Corrected at-large ranges: **51-62** and **98-119**.
  - **Montana**: at-large file covers 51-117; districted 66-102 (no gaps).
    Corrected at-large ranges: **51-65** and **103-117**.
  - **North Dakota**: at-large file covers 51-119; districted 63-72 and
    88-92 (a real gap 73-87, when ND elected two members at-large without
    district lines -- consistent with the "Two general ticket
    representatives" / "One representative" notes already in the data).
    Corrected at-large ranges: **51-62**, **73-87**, and **93-119**.

## Splitting function

```{r}
split_at_large <- function(GEOJSON_FILE, ranges, statefp) {
  gdat <- st_read(GEOJSON_FILE, quiet = TRUE)
  stopifnot(nrow(gdat) == 1, gdat$district == 0)

  split_rows <- map(ranges, function(r) {
    row <- gdat
    row$startcong <- r[1]
    row$endcong <- r[2]
    row$id <- sprintf("%03d%03d%03d%03d", as.integer(statefp), r[1], r[2], 0)
    row
  })

  out <- bind_rows(split_rows)
  unlink(GEOJSON_FILE)
  st_write(out, GEOJSON_FILE, append = FALSE)
  out
}
```

## South Dakota

```{r}
sd_out <- split_at_large(
  file.path(GEOJSON_DIR, "South Dakota_051_to_119.geojson"),
  list(c(51, 62), c(98, 119)),
  statefp = 46
)
st_drop_geometry(sd_out)[, c("statename", "district", "startcong", "endcong", "id")]
```

## Montana

```{r}
mt_out <- split_at_large(
  file.path(GEOJSON_DIR, "Montana_051_to_117.geojson"),
  list(c(51, 65), c(103, 117)),
  statefp = 30
)
st_drop_geometry(mt_out)[, c("statename", "district", "startcong", "endcong", "id")]
```

## North Dakota

```{r}
nd_out <- split_at_large(
  file.path(GEOJSON_DIR, "North Dakota_051_to_119.geojson"),
  list(c(51, 62), c(73, 87), c(93, 119)),
  statefp = 38
)
st_drop_geometry(nd_out)[, c("statename", "district", "startcong", "endcong", "id")]
```

## Validation

For every congress from 51 to 119, and each of the three states, confirm
that no more than one shape from a given file-family now claims that
congress in a way that overlaps another shape for the same congress: i.e.
that the at-large shape and the districted shapes never both claim the
same congress.

```{r}
validate_state <- function(statename) {
  cov <- congress_coverage(statename)
  map_dfr(51:119, function(cng) {
    matches <- cov |> filter(startcong <= cng, endcong >= cng)
    tibble(congress = cng, n_shapes = nrow(matches),
           districts = paste(sort(matches$district), collapse = ","))
  })
}

sd_check <- validate_state("South Dakota")
mt_check <- validate_state("Montana")
nd_check <- validate_state("North Dakota")

# Any congress where an at-large (0) shape coexists with districted (1/2/3)
# shapes would indicate the overlap bug is still present.
bind_rows(
  sd_check |> mutate(state = "South Dakota"),
  mt_check |> mutate(state = "Montana"),
  nd_check |> mutate(state = "North Dakota")
) |>
  filter(str_detect(districts, "0"), str_detect(districts, "1|2|3"))
```

The empty result above confirms that, for every congress 51-119, the
at-large shape and the districted shapes no longer overlap for any of the
three states.

Finally, spot-check the specific congresses called out in the issue
(South Dakota/Montana/North Dakota 92nd; South Dakota 93rd-97th; Montana
93rd-98th and 99th-102nd) to confirm only the correct districted shapes
(no at-large shape) are present:

```{r}
bind_rows(
  sd_check |> mutate(state = "South Dakota"),
  mt_check |> mutate(state = "Montana"),
  nd_check |> mutate(state = "North Dakota")
) |>
  filter(congress %in% c(92, 93, 94, 95, 96, 97, 98, 99, 100, 101, 102))
```
