---
title: "Remove Duplicate Districts From Issue#14"
format: html
editor: source
---

## Overview

@caseydrier pointed out duplicate districts in several geoJson files in 
[Issue #14](https://github.com/JeffreyBLewis/congressional-district-boundaries/issues/14).  Not sure how these survived de-duplication in the processing of the original congress-by-congress files, but I simply remove them here.

## `Alaska_103_to_114.geojson`

```{r}
#| message: false
library(tidyverse)
library(sf)

GEOJSON_FILE <- "../../GeoJson/Alaska_103_to_114.geojson"
gdat <- st_read(GEOJSON_FILE)
```

There are indeed two shapes!

```{r}
glimpse(gdat)
```

They are identical!

```{r}
identical(gdat$geometry[1], gdat$geometry[2])
```

Writing a de-duplicated GEOJSON:

```{r}
unlink(GEOJSON_FILE) 
distinct(gdat) |> 
  glimpse() |>
  st_write(GEOJSON_FILE, append=FALSE)
```


## `California_103_to_107.geojson`

```{r}
GEOJSON_FILE <- "../../GeoJson/California_103_to_107.geojson"
gdat <- st_read(GEOJSON_FILE)
```

There are indeed two distinct CD 13 shapes!

```{r}
gdat |> 
  count(district) |> 
  filter(n > 1)
```

They are the same!

```{r}
dup <- gdat |> filter(district == 13) 
all(st_equals(dup[1], dup[2], sparse = FALSE))
```

```{r}
unlink(GEOJSON_FILE) 
distinct(gdat) |> 
  glimpse() |>
  st_write(GEOJSON_FILE, append=FALSE)
```


## `Maryland_103_to_107.geojson`

```{r}
GEOJSON_FILE <- "../../GeoJson/Maryland_103_to_107.geojson"
gdat <- st_read(GEOJSON_FILE)
```

There are indeed two distinct CD 5 shapes!

```{r}
gdat |> 
  count(district) |> 
  filter(n > 1)
```

They are the same!

```{r}
dup <- gdat |> filter(district == 5) 
all(st_equals(dup[1], dup[2], sparse = FALSE))
```

```{r}
unlink(GEOJSON_FILE) 
distinct(gdat) |> 
  glimpse() |>
  st_write(GEOJSON_FILE, append=FALSE)
```

## `Massachusetts_103_to_107.geojson`

```{r}
GEOJSON_FILE <- "../../GeoJson/Massachusetts_103_to_107.geojson"
gdat <- st_read(GEOJSON_FILE)
```

There are indeed two distinct CD 9 shapes!

```{r}
gdat |> 
  count(district) |> 
  filter(n > 1)
```

They are the same!

```{r}
dup <- gdat |> filter(district == 9) 
all(st_equals(dup[1], dup[2], sparse = FALSE))
```

```{r}
unlink(GEOJSON_FILE) 
distinct(gdat) |> 
  glimpse() |>
  st_write(GEOJSON_FILE, append=FALSE)
```

## `New York_099_to_102.geojson`

```{r}
GEOJSON_FILE <- "../../GeoJson/New York_099_to_102.geojson"
gdat <- st_read(GEOJSON_FILE)
```

There are indeed two distinct CD 29 shapes!

```{r}
gdat |> 
  count(district) |> 
  filter(n > 1)
```

They are the same!

```{r}
dup <- gdat |> filter(district == 19) 
all(st_equals(dup[1], dup[2], sparse = FALSE))
```

```{r}
unlink(GEOJSON_FILE) 
distinct(gdat) |> 
  glimpse() |>
  st_write(GEOJSON_FILE, append=FALSE)
```


## `Washington_103_to_107.geojson`

```{r}
GEOJSON_FILE <- "../../GeoJson/Washington_103_to_107.geojson"
gdat <- st_read(GEOJSON_FILE)
```

There are indeed two distinct CD 9 shapes!

```{r}
gdat |> 
  count(district) |> 
  filter(n > 1)
```

They are the same!

```{r}
dup <- gdat |> filter(district == 3) 
all(st_equals(dup[1], dup[2], sparse = FALSE))
```

```{r}
unlink(GEOJSON_FILE) 
distinct(gdat) |> 
  glimpse() |>
  st_write(GEOJSON_FILE, append=FALSE)
```


## `New York_098_to_098.geojson`, `New York_098_to_102.geojson`, and `New York_099_to_102.geojson`


```{r}
nydat <- bind_rows(
            st_read("../../GeoJson/New York_098_to_098.geojson") |> 
              mutate(fn = "098_to_098"),
            st_read("../../GeoJson/New York_098_to_102.geojson") |> 
              mutate(fn = "098_to_102"),
            st_read("../../GeoJson/New York_099_to_102.geojson") |> 
              mutate(fn = "099_to_102"))
```

`New York_098_to_102.geojson`  Only contains one district, it is artifact of processing and will removed from the collection.

```{r}
nydat |> 
  st_drop_geometry() |>
  count(fn)
```

The question then is do we need separate shapes for the 98th and the 99th through 102nd. The 98th districts are marginally different from the 99th through 102nd:

```{r}
nydat |>
  mutate(area.1 = as.numeric(st_area(geometry))) |>
  st_intersection(nydat) |> 
  glimpse() |>
  filter(fn != fn.1) |> 
  mutate(over_area = as.numeric(st_area(geometry)),
         overlap = over_area/area.1) |> 
  filter(overlap > 0.01,
         fn %in% c("098_to_098"),
         fn.1 %in% c("099_to_102")) |>
  select(fn, fn.1, district, district.1, overlap) |> 
  st_drop_geometry() |>
  arrange(district) |> 
  gt::gt()
```


Here is perhaps the biggest difference:
```{r}
library(osmdata)
library(ggmap)
ny_bb <- nydat |> 
          filter(district == 27) |>
          st_buffer(dist = 1e4) |>
          st_bbox() |> 
          as.matrix() |>
          as.data.frame() |>
          rownames_to_column() |> 
          separate_wider_regex(
            cols = rowname,
            patterns = c(x = "^[xy]",
                         y = ".+")
          ) |> 
  pivot_wider(names_from = y, values_from = V1) |>
  column_to_rownames("x") |>
  as.matrix() 

ny27_map <- get_map(ny_bb, 
                    maptype = "roadmap")
```


```{r}
ggmap(ny27_map) +
  geom_sf(data = nydat |> filter(fn == "098_to_098",
                                 district %in% c(27)),
          inherit.aes = FALSE,
          fill = NA,
          linewidth = 2,
          color = "grey",
          alpha = 0.01) +
  geom_sf(data = nydat |> filter(fn == "099_to_102",
                                 district %in% c(27)),
          inherit.aes = FALSE,
          color = "red",
          fill = NA) +
  ggthemes::theme_map()
```

I need to determine whether there was actual change or whether there was a correction that should have been applied to the 98th was not.  For now, I will leave things as they are.