---
title: "Remove Duplicate Districts From Issue#14"
format: html
editor: source
---

## Overview

@caseydrier pointed out duplicate districts in several geoJson files in 
[Issue #14](https://github.com/JeffreyBLewis/congressional-district-boundaries/issues/14).  Not sure how these survived de-duplication in the processing of the original congress-by-congress files, but I simply remove them here.

## `Alaska_103_to_114.geojson`

```{r}
#| message: false
library(tidyverse)
library(sf)

GEOJSON_FILE <- "../../GeoJson/Alaska_103_to_114.geojson"
gdat <- st_read(GEOJSON_FILE)
```

There are indeed two shapes!

```{r}
glimpse(gdat)
```

They are identical!

```{r}
identical(gdat$geometry[1], gdat$geometry[2])
```

Writing a de-duplicated GEOJSON:

```{r}
unlink(GEOJSON_FILE) 
distinct(gdat) |> 
  glimpse() |>
  st_write(GEOJSON_FILE, append=FALSE)
```


## `California_103_to_107.geojson`

```{r}
GEOJSON_FILE <- "../../GeoJson/California_103_to_107.geojson"
gdat <- st_read(GEOJSON_FILE)
```

There are indeed two distinct CD 13 shapes!

```{r}
gdat |> 
  count(district) |> 
  filter(n > 1)
```

They are the same!

```{r}
dup <- gdat |> filter(district == 13) 
all(st_equals(dup[1], dup[2], sparse = FALSE))
```

```{r}
unlink(GEOJSON_FILE) 
distinct(gdat) |> 
  glimpse() |>
  st_write(GEOJSON_FILE, append=FALSE)
```


## `Maryland_103_to_107.geojson`

```{r}
GEOJSON_FILE <- "../../GeoJson/Maryland_103_to_107.geojson"
gdat <- st_read(GEOJSON_FILE)
```

There are indeed two distinct CD 5 shapes!

```{r}
gdat |> 
  count(district) |> 
  filter(n > 1)
```

They are the same!

```{r}
dup <- gdat |> filter(district == 5) 
all(st_equals(dup[1], dup[2], sparse = FALSE))
```

```{r}
unlink(GEOJSON_FILE) 
distinct(gdat) |> 
  glimpse() |>
  st_write(GEOJSON_FILE, append=FALSE)
```

## `Massachusetts_103_to_107.geojson`

```{r}
GEOJSON_FILE <- "../../GeoJson/Massachusetts_103_to_107.geojson"
gdat <- st_read(GEOJSON_FILE)
```

There are indeed two distinct CD 9 shapes!

```{r}
gdat |> 
  count(district) |> 
  filter(n > 1)
```

They are the same!

```{r}
dup <- gdat |> filter(district == 9) 
all(st_equals(dup[1], dup[2], sparse = FALSE))
```

```{r}
unlink(GEOJSON_FILE) 
distinct(gdat) |> 
  glimpse() |>
  st_write(GEOJSON_FILE, append=FALSE)
```

## `New York_099_to_102.geojson`

```{r}
GEOJSON_FILE <- "../../GeoJson/New York_099_to_102.geojson"
gdat <- st_read(GEOJSON_FILE)
```

There are indeed two distinct CD 29 shapes!

```{r}
gdat |> 
  count(district) |> 
  filter(n > 1)
```

They are the same!

```{r}
dup <- gdat |> filter(district == 19) 
all(st_equals(dup[1], dup[2], sparse = FALSE))
```

```{r}
unlink(GEOJSON_FILE) 
distinct(gdat) |> 
  glimpse() |>
  st_write(GEOJSON_FILE, append=FALSE)
```


## `Washington_103_to_107.geojson`

```{r}
GEOJSON_FILE <- "../../GeoJson/Washington_103_to_107.geojson"
gdat <- st_read(GEOJSON_FILE)
```

There are indeed two distinct CD 9 shapes!

```{r}
gdat |> 
  count(district) |> 
  filter(n > 1)
```

They are the same!

```{r}
dup <- gdat |> filter(district == 3) 
all(st_equals(dup[1], dup[2], sparse = FALSE))
```

```{r}
unlink(GEOJSON_FILE) 
distinct(gdat) |> 
  glimpse() |>
  st_write(GEOJSON_FILE, append=FALSE)
```