---
title: "Fix a Census Cartographic Error in VA CD1/CD3 From Issue #20"
format: html
editor: source
---

## Overview

[Issue #20](https://github.com/JeffreyBLewis/congressional-district-boundaries/issues/20)
reports that Virginia's 103rd-Congress shapefile (`Virginia_103_to_103.geojson`)
disagrees with the 104th/105th-Congress shapefile (`Virginia_104_to_105.geojson`)
along the District 1 / District 3 boundary, in the Middle Peninsula near Essex
County.

Three independent lines of evidence show this is a Census cartographic error in
the 104th/105th-vintage file, not a real mid-decade redistricting:

1. **No enacting law exists.** Former Va. Code &sect; 24.2-302 (the statute
   defining VA's congressional districts) shows no amendment between the 1991
   creation of this map and the court-ordered 1998 redraw. The only citation in
   that window, `Acts 1993, c. 641`, is the wholesale recodification of Title
   24.1 into Title 24.2 -- a renumbering exercise, not a boundary change.
2. **The court record confirms it.** *Moon v. Meadows*, 952 F. Supp. 1141
   (E.D. Va. 1997) -- the case that eventually forced the redraw -- was decided
   in February 1997, after the 104th Congress was already elected, and the new
   map wasn't signed into law until February 1998 (taking effect for the 1998
   election, the 106th Congress).
3. **A period map confirms which side is right.** VPAP's history of the
   district ([vpap.org](https://www.vpap.org/updates/1996-3rd-district-brief-history/))
   includes actual maps of the district as enacted in 1992 and as redrawn in
   1998. The 1992 map has a narrow spike reaching north into the Middle
   Peninsula that is absent from the 1998 map -- i.e., that piece was part of
   District 3 continuously from 1992 through the 1996 election, and was only
   cut loose by the real 1998 redraw, two years after Census's 104th-vintage
   file had already (incorrectly) dropped it.

```{r}
#| message: false
library(tidyverse)
library(sf)
sf_use_s2(FALSE)
```

## Confirming the bug

```{r}
d103 <- st_read("../../GeoJson/Virginia_103_to_103.geojson", quiet = TRUE)
d104 <- st_read("../../GeoJson/Virginia_104_to_105.geojson", quiet = TRUE)

pchng <- function(g1, g2) as.numeric(st_area(st_intersection(g1, g2))) /
  as.numeric(st_area(st_union(g1, g2)))

overlaps <- map_dbl(sort(d103$district), function(dist) {
  g1 <- d103 |> filter(district == dist) |> st_geometry()
  g2 <- d104 |> filter(district == dist) |> st_geometry()
  pchng(g1, g2)
})
tibble(district = sort(d103$district), overlap_103_vs_104 = overlaps)
```

Only Districts 1 and 3 actually differ (overlap well below 1); every other
district is unchanged. This project's own file-writing step groups an entire
state's districts by a shared `(startcong, endcong)` range, so a real change in
even one district splits the whole state's file at that Congress -- which is
why all 11 districts, not just 1 and 3, ended up in a standalone
`Virginia_103_to_103.geojson`.

## Establishing ground truth from Census's own raw historical shapefiles

Both existing files were built from "US Census website" data (see their `note`
field). To check them against an independent source, this pulls Census's own
raw historical congressional district shapefiles directly -- the same
`PREVGENZ` series the repo owner referenced when investigating this issue --
for Virginia (state FIPS 51), the 103rd, 104th, and 105th Congresses:

```{r}
#| message: false
BASE_URL <- "https://www2.census.gov/geo/tiger/PREVGENZ/cd"

load_prevgenz_va <- function(cong) {
  zip_url <- sprintf("%s/cd%dshp/cd51_%d_shp.zip", BASE_URL, cong, cong)
  td <- tempfile()
  dir.create(td)
  zipfn <- file.path(td, "va.zip")
  download.file(zip_url, zipfn, quiet = TRUE)
  unzip(zipfn, exdir = td)
  shpfn <- list.files(td, pattern = "\\.shp$", full.names = TRUE)
  st_read(shpfn, quiet = TRUE) |>
    st_set_crs(4269) |>
    group_by(CD) |>
    summarize(geometry = st_union(geometry), .groups = "drop") |>
    transmute(district = as.numeric(CD), startcong = cong, endcong = cong,
              geometry = geometry)
}

prevgenz <- map(c(103, 104, 105), load_prevgenz_va) |> list_rbind() |> st_as_sf()
```

```{r}
detail <- prevgenz |>
  arrange(district, startcong) |>
  mutate(pchng = map2_dbl(geometry, lag(geometry),
                           \(g1, g2) if (is.null(g2)) NA else pchng(g1, g2)),
         .by = district)
st_drop_geometry(detail)[, c("district", "startcong", "endcong", "pchng")]
```

This independently confirms the same pattern: Districts 1 and 3 change between
103 and 104 (overlap 0.94 and 0.89); every other district is stable across all
three Congresses (104-to-105 overlap is 1.0 everywhere, confirming those two
years are identical). No district changes again between 104 and 105, matching
the fact that the real redraw didn't happen until 1998.

## Locating and checking the disputed area

```{r}
cd3_103 <- prevgenz |> filter(district == 3, startcong == 103) |> st_geometry()
cd3_104 <- prevgenz |> filter(district == 3, startcong == 104) |> st_geometry()

lost <- st_difference(cd3_103, cd3_104) |> st_cast("POLYGON")
areas <- as.numeric(st_area(lost))
main_piece <- lost[[which.max(areas)]]
st_coordinates(st_centroid(st_sfc(main_piece, crs = 4269)))
```

The dominant piece Census's 104th-vintage file drops from District 3 (about
560 million square meters, ~216 sq mi) is centered in Essex County, on the
Middle Peninsula -- exactly the "north to the middle peninsula" leg of
District 3's documented "winding path from Richmond ... to Norfolk"
([VPAP](https://www.vpap.org/updates/1996-3rd-district-brief-history/)), and
exactly where the 1992-vs-1998 map comparison shows the district's boundary
actually changing.

## Fix

Districts 1 and 3 take their 103rd-Congress (accurate) geometry, extended
through the 105th Congress; every other district keeps its existing
104th/105th-Congress geometry (already correct, and closer to current
practice). The result replaces the incorrect three-way split with a single
continuous 103rd-105th Congress file.

```{r}
append_rnote <- function(x, text) {
  x |> mutate(rnote = if_else(
    rnote == "",
    sprintf('{"%s"}', text),
    str_replace(rnote, "\\}$", paste0(',"', text, '"}'))
  ))
}

correction_note <- paste(
  "Fix for GitHub issue #20 (see Patches/11-fix_va_01_03_issue20): Census's",
  "104th/105th-Congress vintage shapefile incorrectly dropped a ~216 sq mi",
  "Middle Peninsula/Essex County piece from this district into District 1.",
  "No Virginia law changed the 3rd District's boundaries between the 1992",
  "and 1998 elections (Moon v. Meadows, 952 F. Supp. 1141 (E.D. Va. 1997),",
  "wasn't decided until Feb. 1997 and the redraw wasn't signed until Feb.",
  "1998); VPAP's 1992 and 1998 district maps confirm this piece was part of",
  "District 3 throughout 1992-1996 and only removed by the real 1998 redraw.",
  "Geometry for Districts 1 and 3 in this range now uses the 103rd-Congress",
  "vintage, which matches the historical record."
)

fixed_1_3 <- d103 |>
  filter(district %in% c(1, 3)) |>
  mutate(startcong = 103, endcong = 105) |>
  append_rnote(correction_note)

fixed_rest <- d104 |>
  filter(!district %in% c(1, 3)) |>
  mutate(startcong = 103, endcong = 105)

new_va <- bind_rows(fixed_1_3, fixed_rest) |>
  mutate(id = sprintf("%03i%03i%03i%03i", 51, startcong, endcong, district)) |>
  arrange(district)
```

## Validate

```{r}
st_drop_geometry(new_va)[, c("district", "startcong", "endcong", "id")]
all(st_is_valid(new_va))

# no overlaps between any two districts
combos <- combn(nrow(new_va), 2)
max(map_dbl(seq_len(ncol(combos)), function(i) {
  a <- new_va[combos[1, i], ] |> st_geometry()
  b <- new_va[combos[2, i], ] |> st_geometry()
  sum(as.numeric(st_area(st_intersection(a, b))))
}))

# total area roughly conserved relative to the 104-105 file it mostly replaces
sum(as.numeric(st_area(new_va)))
sum(as.numeric(st_area(d104)))
```

Eleven districts, no invalid geometry, no pairwise overlap, and total area
essentially unchanged (the Essex County piece simply moves from District 1
back to District 3).

## Write out the corrected file

```{r}
unlink("../../GeoJson/Virginia_103_to_103.geojson")
unlink("../../GeoJson/Virginia_104_to_105.geojson")
st_write(new_va, "../../GeoJson/Virginia_103_to_105.geojson", append = FALSE)
```
